Binding to nucleosome poises SIRT6 for histone H3 de-acetylation

Version 1

Abstract

Abstract

Sirtuin 6 (SIRT6) is a NAD+ dependent histone H3 deacetylase that is prominently found associated with chromatin, attenuates transcriptionally active promoters and regulates DNA repair, metabolic homeostasis and life-span. Unlike other sirtuins it has low affinity to free histone tail but demonstrates strong binding to nucleosomes. It is poorly understood how SIRT6 docking on nucleosomes stimulates its histone de-acetylation activity. Here we present the structure of human SIRT6 bound to a nucleosome determined by cryogenic electron microscopy. The zinc finger domain of SIRT6 associates tightly with the acidic patch of the nucleosome through multiple arginine anchors. The Rossmann fold domain binds to the terminus of the looser DNA half of the nucleosome, detaching two turns of the DNA from the histone octamer and placing the NAD+ binding pocket close to the DNA exit site. This domain shows flexibility with respect to the fixed zinc finger and moves with, but also relative to, the unwrapped DNA terminus. We apply molecular dynamics simulations of the histone tails in the nucleosome to show that in this mode of interaction the active site of SIRT6 is perfectly poised to catalyze deacetylation of the H3 histone tail and that the partial unwrapping of the DNA allows even lysines close to the H3 core to reach the enzyme.

Introduction

The state of lysine acetylation in histones, as well as in several other proteins, is determined by the balancing act of acetyltransferases and deacetylases. Sirtuin is an evolutionarily conserved family of lysine deacetylases that comprises seven members in humans and plays an important role in regulating numerous physiological and pathological processes including aging, cancer, inflammation and metabolism(Cen et al., 2011; Vitiello et al., 2017; Watroba et al., 2017). Unlike others lysine deacetylases sirtuins carry out their enzymatic reaction by transferring the acetyl group to an essential NAD+ co-factor before acetylating the target protein(Imai et al., 2000).

Human Sirtuin 6 (SIRT6) is a multitasking enzyme implicated in diverse pathways. Through its histone H3 deacetylation at gene promoters it is a transcription repressor attenuating inter-alia ribosome biogenesis(Sebastian et al., 2012), glycolysis(Taylor et al., 2022) and the expression of various transcription factors, including nuclear factor κB (NF-κB)(Kawahara et al., 2009) and hypoxia-inducible factor-1α (HIF-1α)(Zhong et al., 2010). Its H3K9 deacetylation of telomeric chromatin maintains telomere integrity(Michishita et al., 2008). In addition, via its second enzymatic activity, ADP-ribosylation of PARP1(Mao et al., 2011) and by stabilizing DNA-PK on chromatin(McCord et al., 2009), SIRT6 also fosters DNA double-strand break repair.

These activities position SIRT6 as a key regulator of longevity as indeed shown by several lines of evidence. In mice, knockout of SIRT6 leads to a shortened lifespan and a premature aging-like phenotype(Kanfi et al., 2012). Similar observations were made in monkeys deficient for the enzyme(Zhang et al., 2018). Conversely, overexpression of SIRT6 extends lifespan and diminishes frailty in mice(Roichman et al., 2021). Furthermore, a study with several rodent species found a positive correlation between SIRT6 ability to promote DNA double-strand break repair and longevity - in other words more active SIRT6 leads to longer lifespan(Tian et al., 2019).

The development of cancer can also be regulated by SIRT6. However, whereas in some malignancies SIRT6 plays a role as a tumor suppressor(Marquardt et al., 2013; Min et al., 2012; Sebastian et al., 2012), in others it acts as an oncogene(Bauer et al., 2012; Khongkow et al., 2013; Liu et al., 2013). The molecular mechanisms underlying SIRT6 effects on longevity or on the two contradictory roles in cancer development are not well understood, limiting the search for medically useful drugs that target the enzyme.

SIRT6 is a 37KDa protein composed of two globular parts, a Rossmann fold domain that harbors the NAD+ binding site and a zinc finger domain(Pan et al., 2011). A striking difference between SIRT6 and other members of the sirtuin family is the presence of a splayed zinc finger and the lack of a stabilizing helix bundle linking the two globular domains. This absence probably endows SIRT6 with a unique flexibility between the domains (Supplemental Figure 1).

These structural differences are manifested in function. Unlike other sirtuins, SIRT6 has a very low activity on acetylated free H3 histones or histone peptides(Jiang et al., 2013). This seemed to contradict findings concerning the function of SIRT6, in particular data demonstrating its activity in vivo as H3 deacetylase(Michishita et al., 2008) as well as the fact that SIRT6 is nuclear and found mainly associated with chromatin(Michishita et al., 2005). The conundrum was settled when it was revealed that nucleosomes, the in vivo cognate substrate of SIRT6, dramatically enhances the enzyme activity towards acetylated H3 histones(Gil et al., 2013). Further biochemical analysis showed that SIRT6 binds nucleosomes with high affinity via multiple interacting sites(Liu et al., 2020).

However, a mechanistic understanding of SIRT6 binding to nucleosome and how this association promotes higher enzyme efficacy is currently lacking.

Here we present the structure of human SIRT6 bound to nucleosome. We identify interactions between the zinc finger domain and the acidic patch of the nucleosome as well as between the Rossmann fold and the DNA terminus, positioning the NAD+ binding pocket close to the nucleosome DNA exit site and partially detaching the DNA from the histone octamer. We use molecular dynamics simulations of the nucleosome to show that in this position the enzyme active site is perfectly poised to bind the H3 tail for deacetylation and that due to the unwrapping of the DNA even lysines close to the core of H3 can reach the enzyme. Furthermore, we find conformational changes, facilitated by the flexibility in SIRT6 between the zinc finger and the Rossmann fold, that could serve to fine-tune the position of the active site with respect to the H3 tail in order to trap its target acetylated residues.

Results

Overall structure

To investigate how nucleosomes stimulate the activity of SIRT6 we prepared nucleosome with the Widom-601 positioning sequence and reconstituted its complex with full-length recombinant human SIRT6 that was overexpressed in and purified from E. coli. The complex was stabilized via mild crosslinking before its deposition on cryo-EM grids.

We obtained a single particle cryogenic electron microscopy structure showing the nucleosome at 2.9 Å overall resolution and a protein density bound to its disc-like surface (Fig. 1a and Supplemental Fig. 2 and 3). SIRT6 could be fitted into this density with its zinc finger domain (resolved at 3.0 Å resolution), associated with the acidic patch of the nucleosome and its Rossmann fold domain (4.0-6.0 Å resolution) interacting with the DNA terminus at superhelix locations (SHL) 6 and 7 (Fig. 1b). The active site of SIRT6 is thus positioned just above where the H3 tail enters the nucleosome disc between the two DNA gyres. The high-resolution at the nucleosome allows discerning purines from pyrimidines. We could therefore distinguish between the two pseudosymmetric orientations of the Widom-601 DNA (Supplemental Fig. 4a). We find that SIRT6 is bound predominantly at the “right” arm of the Widom-601 sequence where interactions with the histone octamer are looser(Ngo et al., 2015). We detect however a small population with a second SIRT6 bound at the “left” tight arm (Supplemental Fig. 2).

Structure of human SIRT6 in complex with the nucleosome

a, Front (top) and side (bottom) views of a composite cryo-EM reconstruction of human SIRT6-nucleosome. Maps from focused refinements of SIRT6 (magenta) and the nucleosome (H2A – yellow; H2B – orange; H3 – blue; H4 – green; DNA – light and dim grey). b, Corresponding views of the atomic model of the complex.

Conformational heterogeneity analyses in cryoDRGN(Zhong et al., 2021) reveals two principle large movements within the structure (Fig. 2a,c).

Flexibility of the SIRT6 Rossmann fold

a, UMAP projection of the latent embeddings of a subset of SIRT6-nucleosome particles representing the concerted movement of the Rossmann fold and the DNA terminus. b, Structural representation of the two endpoints of the latent embeddings shown in panel a. c, UMAP projection of the latent embeddings of another subset of SIRT6-nucleosome particles representing the movement of the Rossmann fold with respect to the nucleosomal DNA. d, Structural representation of the two endpoints of the latent embeddings shown in panel b.

First, we find that where the SIRT6 Rossmann fold binds, DNA is pulled away and detached from the histone octamer (Fig. 2b). Particles greatly vary in the extent of this shift that therefore probably represent a continuous movement of the DNA (compare Fig. 2b1 with 2b2). The Rossmann fold moves with and is attached to the DNA whereas the zinc finger domain remains fixed at the acidic patch of the nucleosome. In the full extent of the DNA movement two turns from SHL 6 and 7 are detached from the histone octamer. This alters the path of the terminus DNA by 37° and breaks DNA interactions with histones H2A, H2B, and H3 (Supplemental Fig. 4b). Such a large change in DNA trajectory is reminiscent of the chromatin-remodeler ATPases or pioneer transcription factors binding to nucleosome but was not observed in other histone modifiers. Importantly, DNA terminus without bound SIRT6 or free nucleosomes in our dataset did not show any deviation from the canonical conformations (Supplemental Fig. 4c). In addition, our MD simulations indicate higher flexibility of the DNA termini when SIRT6 is bound to the nucleosome (Supplemental Fig. 5). All together these findings strongly suggest that the DNA movement we observed is coupled to SIRT6 interactions.

Second, in particles where the SIRT6-bound DNA remains in its canonical position, we observe shifts by up to 20A in the position of the Rossmann fold domain of SIRT6, partially loosening its association with the DNA (Fig. 2d). We show that this shift could be involved in fine tuning the position of the active site relative to the H3 tail.

These conformational changes are made possible by a flexible hinge region between the two domains of SIRT6, which is unique to this member of the sirtuins family.

Zinc finger domain binds the acidic patch

An important feature of the nucleosome surface is a cluster of acidic residues (E56, E61, E64, D90, E91, E92 of H2A and E102, E110 of H2B) that forms a negatively charged ‘acidic patch’ and takes the shape of a narrow groove. It serves as a docking site for many nucleosome binding proteins. Four side chains of the SIRT6 zinc-finger domain extend into different sections of the acidic patch groove and interact with the negatively charged residues. R175 packs against H2B L103 and serves as a classical “arginine anchor motif”(McGinty and Tan, 2016), binding to the acidic triad of E61, D90, E92 from H2A. SIRT6 presents two other arginines, R172, to contact E56 and Q24 of H2A as well as Q44 and E110 of H2B and R178 to contact H2A E91 and E102 from H2B. Finally, K170 interacts with E64 and N68 from H2A (Fig. 3a). All four SIRT6 positive residues are located in a loop that makes a sharp turn, due to a conserved glycine at position 173, connecting two beta strands and penetrates into the acidic patch. Interestingly, this loop appears to be flexible in crystal structures of SIRT6 but rendered stable by its association with the acidic patch. Binding to different sections of the acidic patch maintains a strong hold of the zinc finger at the surface of the nucleosome and orients the Rossmann fold domain towards the DNA terminus. This strong hold is maintained throughout the movements of the Rossmann fold domain (Fig. 2).

Binding of SIRT6 to the nucleosome

a, Close up view on zinc-finger interactions with the acidic patch. Color code as in Fig 1. b, Protein sequence alignment of human sirtuins. Red boxes and asterisks depict the residues of SIRT6 interacting with the acidic patch of the nucleosome. c, Protein sequence alignment of SIRT6 from different species highlighting the same amino acids as in panel b. Organisms: mm Mus musculus, rn Rattus norvegicus, bt Bos taurus, hs Homo sapiens and cc Castor canadensis. d, Depiction of the 3 arginines of SIRT6 (magenta) interacting with the DNA (grey). e, H2A c-terminal tail (yellow) interacts with SIRT6(magenta). f, SIRT6 binding to H2A or H2AZ containing nucleosomes. Bars show the fraction of residual nucleosomes that did not shift with bound SIRT6 in an electron-mobility shift assay (Supplemental Fig. 7). Bars represent mean ± SD of three biological replicates (shown as dots). ** indicates a statistically significant difference between the fraction of residual H2A and H2A.Z containing nucleosomes (P=0.0016 in paired t-test).

It is important to note that the four positive residues of SIRT6 that bind the nucleosome acidic patch are conserved among vertebrates but in contrast are not present in other human sirtuins (Fig. 3b,c). This probably reflects the fact that SIRT6 is the only sirtuin predominantly associated with chromatin.

Rossmann fold domain binds nucleosome DNA exit site

Due to its flexibility with respect to the fixed zinc finger domain, the resolution of the Rossmann fold domain is not sufficiently high to distinguish most individual side chains. However secondary structure elements, mostly helices, are clearly visible and guided the precise fitting of the crystal structure of this domain.

We find that three highly conserved arginines from the Rossmann fold domain (R220, R231, R248) bind the phosphate backbone at SHL 6 and 7 of the nucleosomal DNA (Fig. 3d). These interactions are mostly maintained throughout the movement of the DNA terminus with the Rossmann fold domain bound. However, by comparison to dedicated DNA binding proteins such as transcription factors or TBP the number of observed DNA-protein interactions is low in SIRT6. A biochemical study suggests that the intrinsically disordered C-terminal domain of SIRT6 is necessary for establishing tight interaction of SIRT6 with the nucleosome(Liu et al., 2020) via its ability to bind DNA. Our maps do show an additional isolated protein density that we could not assign with confidence to SIRT6 but associates with the DNA terminus Supplemental Fig. 6a). We propose however that the three arginines are required to correctly position the Rossmann fold, and hence the catalytic site, with respect to the H3 tail.

Unlike other sirtuins, SIRT6 binds NAD+ in the absence of an acetylated substrate. This is attributed to the fact that crystal structures of SIRT6 describe a rigid helix (α3) where other sirtuins contain a flexible cofactor binding loop(Pan et al., 2011). In our map we do not observe helix α3 (Supplemental Figure 6b) and we speculate that in the context of the nucleosome this helix assumes a flexible loop conformation that like in other sirtuins will bind the cofactor only after stabilization by the acetyl lysine substrate.

In addition, we find that the N-terminal part of the Rossmann fold domain associates with the C-terminal tail of histone H2A at residue K119 (Fig. 3e). This observation suggests that monoubiquitination of H2A at this position, a major histone modification in mammalian cells, could interfere with the binding of SIRT6 or regulate its activity. Furthermore, we used electrophoretic mobility shift assay to follow the binding of SIRT6 to nucleosomes harboring the histone variant H2A.Z which deviates considerably from the canonical H2A in its C-terminal tail but not in other residues that contact SIRT6. Surprisingly, we find that binding of SIRT6 to nucleosomes is mildly (but statistically significant) stronger when H2A.Z is incorporated instead of H2A (Fig. 3f and Supplemental Fig. 7). This finding could be related to SIRT6 role in repressing gene expression because H2A.Z is enriched in nucleosomes that flank the transcription start site.

SIRT6 is poised to deacetylate the H3 tail

The classical model of the nucleosome posits an elongated and disordered histone H3 N-terminal tail that freely roams and extends away from the core. However recent NMR studies and molecular dynamics predictions show that the tail actually collapses onto the nucleosome core, driven by robust interactions with DNA(Ikebe et al., 2016; Stutzer et al., 2016). The tail adopts multiple structurally heterogeneous, but energetically similar, states. It dynamically transitions between these conformations that all interact with the last superhelix location of nucleosomal DNA(Morrison et al., 2018). We used a published set of 15 predicted H3-tail conformations resulting from end-states of molecular dynamics simulations of the whole nucleosome(Morrison et al., 2018), to analyze whether the position of SIRT6 on the nucleosome serves the enzyme’s role in trapping residues of histone H3 tail. The nucleosome from each molecular dynamics simulation was superposed on our model of SIRT6 bound to a nucleosome with the DNA terminus at the canonical position. To represent the full range of movement of the Rossmann fold domain relative to the DNA we used two endpoints of this movement namely fully bound to DNA and partially dissociated from it (Fig. 4a and b, respectively). We then examined the position of the SIRT6 active site with respect to the predicted paths of the histone H3 tail.

SIRT6 poised to deacetylate lysine residues of H3

a, Side View of SIRT6 bound to the nucleosome with histone H3 K9 residue (red sphere) closest to the SIRT6 active site (blue spheres) from a set of molecular dynamics simulations. b, Side View of SIRT6 bound to the nucleosome with histone H3 K18 residue (orange sphere) closest to the SIRT6 active site from a set of molecular dynamics simulations. c, All H3K9 positions (red spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. d, All H3K18 positions (orange spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. e, Molecular dynamics simulations show that H3 c-terminal tail (blue) can protrude towards SIRT6 in a space formed between the histone octamer and the DNA. H3K27 is shown in red.

We find that in several H3 conformations the tail is located in close proximity (<15Å) to the SIRT6 active site. In some of these cases, H3K9 or H3K18, the main target residues of SIRT6(Wang et al., 2016), are practically situated in the active site. Furthermore, when analyzing the two endpoints of the Rossmann fold domain shift with respect to the DNA, we observe that in some H3 conformations this shift draws the active site closer to the target residues (Fig. 4c,d). We conclude that the high-affinity binding of SIRT6 to the nucleosome poises the active site to trapping its main target residues. In addition, the flexibility of the Rossmann fold domain serves as a fine-tuning mechanism that allows for short-range scanning of the tail in order to locate the acetylated residues. We note also that except H3 no other histone is present at the vicinity of the SIRT6 active site. This gives rise to a selectivity of SIRT6 towards the H3 tail.

To shed light on possible benefits offered by SIRT6-mediated DNA unwrapping we conducted molecular dynamic simulations of a nucleosome with bound SIRT6. Strikingly, new H3 tail conformations appear in the simulations that are absent from a control run with nucleosome alone. These pathways have the H3 tail protruding out of the histone core through the space between the DNA and the octamer that is formed via partial unwrapping of the DNA by SIRT6. Importantly, in these additional conformations H3 tail does not need to circle around the DNA to reach SIRT6 and therefore lysines that are closer to the core of H3 (e.g., K27) could also be accessible to the enzyme (Fig. 4e).

Discussion

Most histone modifying enzymes, including sirtuin family members such as SIRT1, show significant activity on isolated target peptides and form multiple interactions with the residue to be modified as well as with flanking amino acids to provide selectivity. It is therefore striking that the measured turnover rate for SIRT6 on isolated H3K9 peptides is 1000 slower than for other sirtuins. Moreover, unlike SIRT1, the activity of SIRT6 is dramatically stimulated by binding to its cognate nucleosome substrate.

We solved the structure of SIRT6 bound to a nucleosome and used heterogeneity analysis to uncover the main conformational changes in the complex. Examining these findings in the context of molecular dynamics studies of the nucleosome that produce putative paths of histone tails, shed light on the unique properties of SIRT6.

We find that multiple interactions of SIRT6 with the nucleosome position its active site at the putative path of relatively long-lived conformations of H3 so that residues K9 and K18 of the tail spend significant time very close to, or even at, the enzyme active site. In addition, SIRT6 moves with respect to the nucleosomal DNA allowing it to finetune the position of the active site so as to trap the target residue. Hence it is the binding of SIRT6 to the nucleosome and an intrinsic scanning capability that account for its much higher enzymatic efficiency on nucleosomes. Importantly, the position of SIRT6 also ensures high selectivity because the space surrounding the active site is devoid of any other tails apart from H3.

The recognition mode of nucleosomal H3 employed by SIRT6 is reminiscent of that used by the PRC1 complex in the specific ubiquitylation of H2A Lys119(McGinty et al., 2014). Similar to SIRT6, PRC1 requires a nucleosome substrate, showing no activity on an H2A-H2B dimer, and multiple interactions with the nucleosome position an active site in close proximity to its target residue. However, the flexibility and scanning capacity we observed for SIRT6 is lacking in PRC1 perhaps echoing the fact that its target residue is rather stable in the nucleosome compared to the H3 tail.

It is interesting to note that the flexibility of SIRT6 that is at the root of its scanning mechanism was suggested to be, at least in part, responsible for its nearly undetected activity on isolated peptides(Pan et al., 2011). It is only when it is constrained to the proximity of the target residues, through binding to the nucleosome, that this capacity facilitates the function of the enzyme.

A second major conformational change observed in our 3D reconstructions is the detachment of the DNA terminus, with a bound Rossmann fold domain, from the octamer. Our molecular dynamics simulations suggest that the space formed between DNA and octamer allows for additional conformations of H3 tail that protrudes towards SIRT6. These new pathways do not require circling around the DNA gyre and therefore lysines that are closer to the core of H3 (e.g., K27) can reach in this way the active site of SIRT6. We note that K56 which lies at the core of H3 within a stable helix was also reported to be a target for SIRT6. It is therefore possible that the enzyme employs additional recognition modes of its targets.

Because the position of SIRT6 active site seems to be well suited for modifying H3 tail, it stands to reason that other H3 modifying enzymes, such as the GCN5 HAT module of SAGA and ATAC, would place their active site in a position similar to the one of SIRT6 and possibly also displace the DNA from the histone octamer. Gcn5 deviates from SIRT6 in having significant affinity for the H3 tail and therefore might pose less stringent requirement for precise positioning of the active site or for specific conformations of the H3 tail. Thus, we speculate the Gcn5 recognition mode will be similar to SIRT6 but will also include some sequence specific binding.

Methods

Human SIRT6 purification

Plasmid (AddGene #28271) was transformed into E. coli strain BL21-RIL. Cells in 1L TB (with 50 µg/ml Kanamycin and 25 µg/ml Chloramphenicol) grew at 37C degrees till OD260 reached 0.8 when they were rapidly cooled on ice to 18C. Over-expression was induced by the addition 0.5 mM IPTG and cells continued to grow at 18C over-night. All subsequent steps were performed at 0-4C and solutions were always supplemented with a protease inhibitor cocktail (PMSF, Peptstatin, E-64). Cells were harvested by centrifugation at 8250g for 20 minutes, suspended in buffer A200 (20 mM Hepes pH 7.5, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 200 mM NaCl) and treated with lysosyme (1 mg/ml) for 30 minutes on ice. Salt concentration was increased to 0.5M and cells were then lysed by sonication (6*10sec). Debris were pelleted by centrifugation 45,000g*20minutes. The supernatant was incubated with 0.5 ml of Ni beads (1 ml slurry, Roche), equilibrated in buffer A500, for 1h. The solution was poured into a gravity column (CliniScience). Beads were then washed by 10 ml of buffer A500 with 5mM imidazole before elution by 2ml of buffer B300 (20 mM Hepes pH 7.2, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 300mM NaCl) with 150mM imidazole. The eluted protein was diluted with two volumes of buffer D (20 mM Hepes pH 7.0, 5% glycerol v/v, 2 mM beta-mercaptoethanol), loaded on a 1ml Heparin column and eluted with 0.1M-1M salt gradient in buffer D. Fraction were analyzed by SDS-PAGE and those containing SIRT6 were pooled and dialyzed against buffer E (20 mM Hepes pH 7.5, 2% glycerol v/v, 0.5 mM TCEP and 150 mM NaCl. Human SIRT6 was then concentrated to 7.8 mg/ml, flash frozen in N2, and kept in aliquotes at -80C.

Octamer reconstitution and nucleosome DNA

Octamers were reconstituted from individual Xenopus leavis (canonical) or human (H2A.Z) histones expressed as inclusion bodies according to the standard protocol(Dyer et al., 2004; Luger et al., 1999). Widom 601 145bp DNA was produced using a plasmid harboring 16 copies of this sequence as described by Dyer et al.(Mattiroli et al., 2017).

Nucleosome core particle reconstitution

Nucleosomes with 145bp Widom 601 positioning sequence were prepared according to NEB Dilution Assembly Protocol (E5350) (https://international.neb.com/protocols/2012/06/02/dilution-assembly-protocol-e5350) with some modifications as follows: 2.75 µM 145 bp Widom601 DNA was mixed with 2.75 µM canonical or H2AZ histone octamers in a solution containing 2M NaCl, 1mM EDTA, 5mM beta-mercaptoethanol. The solution was incubated for 30 min at RT and then underwent serial dilutions down to 1.48M, 1M, 0.6M, 0.25M NaCl with buffer LowSalt (10 mM Hepes-KOH pH 8.0, 2.5 mM beta-mercaptoethanol). After each dilution the solution was kept at RT for 30 min. In order to reduce the final NaCl concentration, nucleosomes were concentrated in 0.5 mL 100KDa cutoff Amicon up to 100 µl, then diluted 5 times with buffer LowSalt. This step was repeated one more time. Finally, nucleosomes were concentrated to 3-4 µM and analyzed in a 5% native 0.2xTBE polyacrylamide gel to ascertain the quality of the sample and absence of free DNA.

Binding and gel-shift assays

For nucleosome-binding assays, 100 nM of recombinant nucleosomes (harboring either H2A or H2AZ histone) were incubated in binding buffer (10 mM Hepes, 50 mM NaCl, 0.5 mM TCEP, 6% glycerol, and pH 7.5) with SIRT6 at 1:2 molar ratio. The reactions were performed on ice for 1h in LoBind tubes (Eppendorf) to minimize protein loss. The reactions were then resolved in a 0.2x TBE 5% gel. The gel was stained with Ethidium Bromide. Bands were quantified with ImageJ.

Grid preparation

SIRT6 was first diluted in buffer (20mM Hepes-K pH 7.5, 150mM NaCl, 0.5 mM TCEP, 2% Trehalose). Nucleosome and SIRT6 were then mixed at a 1:2 molar ratio. Final conditions included 1.7 µM Nucleosome and 3.4 µM SIRT6 in buffer (20mM Hepes-K pH 7.5, 18mM NaCl, 0.5mM TCEP). After 1h on ice glutaraldehyde and the detergent dodecyl-maltoside were added at a final concentration of 0.05% v/v and 0.0125% w/v, respectively. The sample was incubated with the cross-linker for an additional 30-60 minutes. 2nm carbon foil was floated on 3.5/1 300 mesh Quantifoil grids. 3 µl of sample was deposited onto such grids, blotted 5s with blot force 5 after 60s incubation in a Vitrobot IV (Thermo Fisher) at 6 °C using 100 % humidity and flash-frozen in liquid ethane.

cryo-EM data acquisition

Two different datasets were collected. For the first dataset images were acquired on a Cs-corrected Titan Krios G1 (Thermo Fisher) microscope operating at 300 kV in nanoprobe mode using serialEM for automated data collection. Movie frames were recorded on a Gatan K3 direct electron detector after a Quantum LS 967 energy filter using a 20 e-V slit width in zero-loss mode. Images were acquired hardware-binned at a nominal magnification of 130,000, which yielded a pixel size of 0.862 Å. Forty movie frames were recorded at a dose of 1.12 and 1.30 e− per Å2 per frame. The second image dataset was acquired on a Titan Krios G4 microscope operating at 300 kV in nanoprobe mode using EPU for automated data collection. Movie frames were recorded on a Flacon 4i direct electron detector after a Selectris X energy filter using a 10 e-V slit width in zero-loss mode. Images were acquired at nominal magnification of 270,000, which yielded a pixel size of 0.458 Å.

Image analysis

For the first dataset movie frames were aligned, dose-weighted in Warp(Tegunov and Cramer, 2019) together with CTF estimation and particles picking. For the second dataset frame alignment, dose-weighting and CTF estimation was done in cryoSPARC(Punjani et al., 2017), picking was done with crYOLO(Wagner et al., 2019). These datasets were analyzed in cryoSPARC and RELION 3.1(Zivanov et al., 2018) according to standard protocols. Briefly, images were subjected to reference-free 2D classification in cryoSPARC to remove images with contaminations. Ab-initio reconstruction was performed on the selected images followed by heterogenous and homogenous refinements in cryoSPARC. Structural variability and flexibility were explored with Relion 3.1 using 3D classification without alignment with different T values and cryoDRGN(Zhong et al., 2021) according to standard protocols. Sharpening of the maps was done with DeepEmhancer(Sanchez-Garcia et al., 2021).

Model building and refinement

Crystal structure of Widom 601 sequence containing nucleosome (PDB: 3LZ0) and human SIRT6 (PDB: 5×16) were rigid-body fitted into the cryo-EM structures in Chimera(Pettersen et al., 2004). Flexible-fittings were performed with Isolde(Croll, 2018) in ChimeraX(Goddard et al., 2018). The models were refined in real-space with PHENIX(Liebschner et al., 2019) and Isolde.

Molecular Dynamics simulations

All molecular dynamics (MD) simulations were performed with the AMBER20 suit of programs, using the ff14SB(Maier et al., 2015) force field with bsc1 DNA corrections(Ivani et al., 2016) and CUFIX ions corrections(Yoo and Aksimentiev, 2018) that improve the description of disordered region conformations. Parameters for the zinc finger were taken from Macchiagodena et al.(Macchiagodena et al., 2020).

The histone tails were added to the cryo-EM structure based on the work described by Armeev et al(Armeev et al., 2021), and protonation states of the protein residues were checked with the propka3 software(Olsson et al., 2011). The system was soaked in a truncated octrahedral TIP3P water box of 20Å buffer and 0.15 mM of NaCl ions were added to simulate physiological conditions, resulting in a total of 330,000 atoms.

The minimization of the energy was carried out in 5 steps, each one alternated with a 20ps NVT equilibration step at 100K to ensure a very smooth optimization of the starting geometry. Each minimization run included 5,000 steps in steepest descent and 5,000 steps in conjugate gradient, with decreasing restraints on the protein and nucleic atoms from 20kcal/mol to 5kcal/mol that were kept in the associated equilibration step. The last minimization step was released of any constraints and followed by a final NVT equilibration run at 300K. The time step was then switched from 2 to 4fs as allowed by the use of the SHAKE bond length constraints(Kräutler et al., 2001) and the Hydrogen Mass Repartioning algorithm(Hopkins et al., 2015). The latter was applied on the topology using the parmed tool of Amber20. A 40ns equilibration run in NTP was then carried out, followed by a 5 µs production. Coordinates were outputted every 1ns during the production run. The temperature was kept constant using the Langevin thermostat with a 2ps-1 collision frequency, and the Berendsen barostat was used to maintain the pressure at 1 atm in the NTP runs. Electrostatic interactions were treated using the Particle Mesh Ewald approach(Darden et al., 1993) with an 8Å cut-off. Three replicates were performed starting with different velocities at the unconstrained 300K NVT equilibration step, resulting in a total of 15 µs of sampling.

A cluster analysis was performed on the MD ensembles based on the DNA and protein RMSD. Conformations were extracted every 5 frames of the trajectories to avoid memory overload, and were sorted using the hierarchical agglomerative and average linkage algorithms. 10 clusters were generated by replicate (Supplemental Table 2).

The DNA flexibility analysis was performed on each MD ensemble by using a Principal Component Analysis-based machine learning script quantifying the contribution of each residue to the overall dynamics of the system. Based on the Scikit-learn58 library(Fleetwood et al., 2020), it calculates a covariance matrix from the internal coordinates of the MD ensemble. In this matrix, the eigenmodes showing the highest amplitude are the most implicated in the main fluctuations of the system, and the contribution of each residue is normalized based on these values.

The control system was generated by removing SIRT6 from the initial structure and was submitted to the same protocol.

Supporting information

eLife Assessment

This manuscript provides a useful reconstruction of the structure of the sirtuin-class histone deacetylase Sirt6 bound to a nucleosome based on cryo-EM observations, and additional characterization of the flexibility of Sirt6 based on molecular dynamics simulations. The analysis of the cryo-EM data is incomplete for some of the conclusions, and certain elements of the presentation were inadequate to allow sufficient evaluation. Biochemical validation of the conclusions is not provided, but some of this evidence has been published recently by two other groups in their analyses of the same complex. While much in this manuscript is confirmatory, the work also includes new insights into the potential dynamics of Sirt6 bound to a nucleosome.

Joint Public Review:

Smirnova et al. present a cryo-EM structure of human SIRT6 bound to a nucleosome as well as the results from molecular dynamics simulations. The results show that the combined conformational flexibilities of SIRT6 and the N-terminal tail of histone H3 limit the residues with access to the active site, partially explaining the substrate specificity of this sirtuin-class histone deacetylase. The cryo-EM analysis in its current form is incomplete, lacking aspects of validation such as angular distribution information and other standard measurements of the quality of the reconstruction. Biochemical validation of the structural findings is inadequate, relying primarily on previous publications. Importantly, two other groups have recently published cryo-EM structures of SIRT6:nucleosome complexes. This manuscript by Smirnova et al., therefore, confirms and complements these previous findings, with the addition of some novel insights into the role of structural flexibility in substrate selection.

Author Response

The following is the authors’ response to the original reviews.

We are grateful to the reviewers for their remarks which significantly improved the paper. Following these remarks we completed the analysis and validation of our cryo-EM data and peformed several biochemical tests to support our conclusions, lending credbility to the paper. Please find our detailed answers bellow each recommendation of the reviewers.

Major recommendations

1. Errors and omissions in the presentation make the manuscript difficult to access.

a) The text should be edited for grammatical errors more carefully

We corrected the grammatical errors.

b) Figures should be labeled to allow the reader to follow the logic of the presentation and identify the features being discussed. Identification through the color coding (the identity of the histones, the location of zinc fingers, the active site, and so on) would be helpful.

We labeled the Rossman fold and Zn-finger domains in Figure 1 and described the histone color codes. The active site of SIRT6 is depicted in Figure 4.

1. The recent publications from the Farnung/Cole and Peterson/Tan/Armache labs need to be cited and the results from Smirnova et al. compared and contrasted with those publications explicitly.

We added the following paragraph to the discussion section:

“While this manuscript was under review two studies describing the structure of SIRT6-NCP appeared in press (Wang et al., 2023 ; Chio et al., 2023). The conclusion of these papers regarding the position of SIRT6 on the nucleosome and the unwinding of DNA by the enzyme are similar to our findings. We however dissected in addition the movements of SIRT6 on the nucleosome and analyzed via molecular dynamics the conformations of the H3 tail with respect to the SIRT6 active site. Our results point to the importance of the flexibility between the globular domains of SIRT6 and also explain how SIRT6 can access lysines that are much closer to the histone core than H3K9.”

a) Notably, the Peterson/Tan/Armache labs suggest that H3K27 cannot be deacetylated by SIRT6 whereas the Farnung/Cole labs show deacetylation of H3K27 by SIRT6. Do the results of the Smirnova et al. structure help to resolve this situation?

We performed deacetylation tests of H3K27Ac nucleosomes and show that SIRT6 deacetylate H3K27Ac albeit at somewhat lower efficiency than H3K9Ac. Our molecular dynamics simulations explain how H3K27, which is close to the histone core, can still be reached by SIRT6 active site. We added the following text to the paper: “To lend support to this claim we tested whether SIRT6 can deacetylate residue H3K27 that was first acetylated by SAGA (Supplemental Fig. 7c). We find that indeed SIRT6 could efficiently deactylate H3K27Ac, although at a somewhat slower rate than H3K9Ac. We conclude that partial DNA unwrapping by SIRT6 allows H3-tail conformations that make lysines that are close to the core of H3 accessible to the enzyme.”

b) The Farnung/Cole labs have visualized an intermediate state of deacetylation. How does this compare to the structure presented in this manuscript? Addressing these points would facilitate further research and discussion in the community.

We believe the resolution of the SIRT6 Rossmann fold precludes addressing these points.

c) Can the authors exclude the possibility that the additional density observed in Supplemental Figure 6 is not coming from the H3 tail, as observed in the two other structures?

One density is the continuation of the H2A histone tail. We strongly believe that this density corresponds to this tail. The other density indeed can originate from the H3 tail. Therefore, we didn’t model anything inside it.

d) It would be useful to comment on how much flexibility has been observed in the other structures for the SIRT6 interaction with the acidic patch, and also how other acidic-patch binding proteins compare with the results here.

We refrain from estimating the flexibility observed in the other structures as no such analysis is provided by these papers. Regarding the interaction with the acidic patch we mention that R175 packs against H2B L103 and serves as a classical “arginine anchor motif” and refer the reader to a review on the topic.

e) Does the presence or absence of NAD+ affect the comparisons among the structures?

NAD+ binding might affect the fine structure of the active site although NAD+ was not observed in crystal stuctures of SIRT6 in its presence. The resolution of this part precludes further addressing this issue.

1. The lack of biochemical validation of conclusions should be acknowledged and the reasoning behind this choice discussed.

We added experiments to validate our conclusions with biochemical tests. We produced nucleosomes with acetylatexd histone H3 by employing purified SAGA acetyltransferase complex. We isolated SIRT6 where the four residues implicated in interactions with the acidic patch are mutated to alanines (SIRT6-4A). We show that this mutant has very weak interaction with the nucleosome and much lower H3K9Ac deacetylation activity than WT. Similarly SIRT6-3A with mutations in the residues we suggest involved in binding to nucleosomal DNA also shows weak activity and binding to the nucleosome. We added Supplement Figure 7 that depicts the results of these experiments and embedded reference to these results in the approporiate sections of the text. Furthermore, we also show that SIRT6 is active in deacetylating H3K27Ac. This supports our molecular dynamics simulations showing that when SIRT6 binds the nucleosome, H3 tail can assume conformations where H3K27 is accessible by the enzyme’s active site. These results also appear in Supplement Figure 7.

1. The authors nicely analyze and discuss the conformational flexibility of SIRT6 binding. This is an interesting finding, but Fig. 2 does not adequately convey this flexibility.

We now considerably improved Figure 2. We added panels c and f which depict clearly the movements we observe.

1. The authors need to explain why two cryo-EM datasets were collected but were not merged, and the labeling of the datasets in the Supplemental Table appear to be switched.

The two datasets were collected with two very different pixel spacing therefore merging the two was possible only in Relion. This process, however, did not improve the resolution of the SIRT6’s Rossmann fold domain. We thank the reviewer to notice the discrepancy in the text and the Supplemental Table 1, it was corrected.

1. Supplemental Figure 4 should be expanded to show additional representative densities with the respective fit of the model. This will allow the reader to better judge the quality of the data. At least the acidic patch interaction, the DNA-SIRT6 interactions, and the H2A should be shown in this context.

To illustrate the high-resolution features of the structure as well as the key regions we added Supplemental Figure 4.

1. Standard elements of data analysis and validation should be included (angular distribution plots for cryo-EM reconstructions, a 3D FSC sphericity plot, a Q-score and EMRinger score for the cryo-EM data and atomic model, a model-to-map FSC curve). In general, model building is poorly described as it is unclear which maps (or to what degree different maps) were used for this process. This should be clarified in the methods section and in the Supplemental Table 1.

The model validation and data analysis details were added to Supplemental Figures 2 and 3 as well as in Supplemental Table 1.

1. The provided maps also do not fully recapitulate the path of the H2A tail. The various density maps and PDB provided for this review do not support the final modeled residues of H2A between residues #118/119-123. This affects the validity of figure 3E and the discussion of the proximity of the potential substrates to the active site. The authors should clarify how they inferred that this is the H2A tail rather than the loosely bound SIRT6 Nterminal loop (whose stability could be altered by the presence or absence of NAD+) as suggested by overlaying the relevant crystal structures.

We added a panel to Supplemental Figure 4 (d) depicting the density where the H2A tail was modelled.

1. The authors should explain how the data produced an asymmetrically oriented complex with a single SIRT6 molecule bound to one face. Were complexes with two SIRT6 molecules excluded? Is supplementary figure 4A the basis for the orientation and is this sufficient for this purpose?

Complexes with two SIRT6 molecules were present but only at around 1.5 percent of the whole dataset. These images were excluded from the refinement (shown in Supplementary Figure 2). The DNA orientation is depicted in Supplementary Figure 5A. The resolution obtained at the dyad (~2.5Å) allowed us to distinguish purine and pyrimidine bases. The Widom 601 sequence is asymmetric and the densities clearly show that there is only one orientation of the DNA observed with respect to SIRT6.

1. The authors should clarify how supplemental figure 4B supports the conclusion that DNA is unwrapped. The density is not readily visible and docking of a simple DNA model in the ZN-focused map does not clearly rule out the possibility that this density comes from the H3 N-terminal tail.

We added to this figure the cryo-EM densities used to model the DNA path and the orientation of SIRT6. This image is now Supplemental Figure 5c.

Minor recommendations

1. The scale bar is missing for the 2D classes shown in Supplemental Figure 2.

We added the scale bar to the image depicting the 2D classes in Supplemental Figure 2.

1. Masked classifications should be shown in the classification tree (Supplemental Figure 2 +3) with the masks shown as a transparent volume.

We now show the mask used for the 3D classifications of the SIRT6’s Rossman fold domain in Supplemental Figure 2.

1. Supplemental Figure 3 should show the indicated 3D classifications in the classification tree.

We added the 3D classifications in Supplemental Figure 3.

1. The authors should consider applying local CTF refinement and particle polishing to improve their resolution.

We did local and global CTF refinements. Polishing didn’t improve the resolution as movie frame alignment was done outside of Relion.

1. The descriptions of the Widome 601 sequence orientation should be less ambiguous, perhaps mentioning the AT-rich and AT-poor arms instead of left and right arms.

We corrected the text as required.

1. The statement "Such a large change in DNA trajectory is reminiscent of the chromatin-remodeler ATPases or pioneer transcription factors binding to nucleosome but was not observed in other histone modifiers" requires a citation.

We added approporiate references.

1. The authors should provide a supplemental figure of the nucleosome-SIRT6 and PRC1-nucleosome structure comparison to complement the discussion section.

We refer the reader to the paper describing the PRC1-nucleosome structure.

Version 2

Abstract

Abstract

Sirtuin 6 (SIRT6) is an NAD+ dependent histone H3 deacetylase that is prominently found associated with chromatin, attenuates transcriptionally active promoters and regulates DNA repair, metabolic homeostasis and lifespan. Unlike other sirtuins, it has low affinity to free histone tail but demonstrates strong binding to nucleosomes. It is poorly understood how SIRT6 docking on nucleosomes stimulates its histone deacetylation activity. Here we present the structure of human SIRT6 bound to a nucleosome determined by cryogenic electron microscopy. The zinc finger domain of SIRT6 associates tightly with the acidic patch of the nucleosome through multiple arginine anchors. The Rossmann fold domain binds to the terminus of the looser DNA half of the nucleosome, detaching two turns of the DNA from the histone octamer and placing the NAD+ binding pocket close to the DNA exit site. This domain shows flexibility with respect to the fixed zinc finger and moves with, but also relative to, the unwrapped DNA terminus. We apply molecular dynamics simulations of the histone tails in the nucleosome to show that in this mode of interaction, the active site of SIRT6 is perfectly poised to catalyze deacetylation of the H3 histone tail and that the partial unwrapping of the DNA allows even lysines close to the H3 core to reach the enzyme.

Introduction

The state of lysine acetylation in histones, as well as in several other proteins, is determined by the balancing act of acetyltransferases and deacetylases. Sirtuin is an evolutionarily conserved family of lysine deacetylases that comprises seven members in humans and plays an important role in regulating numerous physiological and pathological processes including aging, cancer, inflammation and metabolism (Cen et al., 2011; Vitiello et al., 2017; Watroba et al., 2017). Unlike other lysine deacetylases sirtuins carry out their enzymatic reaction by transferring the acetyl group to an essential NAD+ cofactor before acetylating the target protein (Imai et al., 2000).

Human Sirtuin 6 (SIRT6) is a multitasking enzyme implicated in diverse pathways. Through its histone H3 deacetylation at gene promoters, it is a transcription repressor attenuating inter-alia ribosome biogenesis (Sebastian et al., 2012), glycolysis (Taylor et al., 2022) and the expression of various transcription factors, including nuclear factor κB (NF-κB) (Kawahara et al., 2009) and hypoxia-inducible factor-1α (HIF-1α) (Zhong et al., 2010). Its H3K9 deacetylation of telomericchromatin maintains telomere integrity (Michishita et al., 2008). In addition, via its second enzymatic activity, ADP-ribosylation of PARP1 (Mao et al., 2011) and by stabilizing DNA-PK on chromatin (McCord et al., 2009), SIRT6 also fosters DNA double-strand break repair.

These activities position SIRT6 as a key regulator of longevity as indeed shown by several lines of evidence. In mice, knockout of SIRT6 leads to a shortened lifespan and a premature aging-like phenotype (Kanfi et al., 2012). Similar observations were made in monkeys deficient for the enzyme (Zhang et al., 2018). Conversely, overexpression of SIRT6 extends lifespan and diminishes frailty in mice (Roichman et al., 2021). Furthermore, a study with several rodent species found a positive correlation between SIRT6 ability to promote DNA double-strand break repair and longevity -in other words, more active SIRT6 leads to longer lifespan (Tian et al., 2019).

The development of cancer can also be regulated by SIRT6. However, whereas in some malignancies SIRT6 plays a role as a tumor suppressor (Marquardt et al., 2013; Min et al., 2012; Sebastian et al., 2012), in others it acts as an oncogene (Bauer et al., 2012; Khongkow et al., 2013; Liu et al., 2013). The molecular mechanisms underlying SIRT6 effects on longevity or on the two contradictory roles in cancer development are not well understood, limiting the search for medically useful drugs that target the enzyme.

SIRT6 is a 37 KDa protein composed of two globular parts, a Rossmann fold domain that harbors the NAD+ binding site and a zinc finger domain (Pan et al., 2011). A striking difference between SIRT6 and other members of the sirtuin family is the presence of a splayed zinc finger and the lack of a stabilizing helix bundle linking the two globular domains. This absence probably endows SIRT6 with a unique flexibility between the domains (Supplemental Figure 1).

These structural differences are manifested in function. Unlike other sirtuins, SIRT6 has very low activity on acetylated free H3 histones or histone peptides (Jiang et al., 2013). This seemed to contradict findings concerning the function of SIRT6, in particular data demonstrating its activity in vivo as H3 deacetylase (Michishita et al., 2008) as well as the fact that SIRT6 is nuclear and found mainly associated with chromatin (Michishita et al., 2005). The conundrum was settled when it was revealed that nucleosomes, the in vivo cognate substrate of SIRT6, dramatically enhance the enzyme activity towards acetylated H3 histones (Gil et al., 2013). Further biochemical analysis showed that SIRT6 binds nucleosomes with high affinity via multiple interacting sites (Liu et al., 2020).

However, a mechanistic understanding of SIRT6 binding to nucleosomes and how this association promotes higher enzyme efficacy is currently lacking.

Here we present the structure of human SIRT6 bound to the nucleosome. We identify interactions between the zinc finger domain and the acidic patch of the nucleosome as well as between the Rossmann fold and the DNA terminus, positioning the NAD+ binding pocket close to the nucleosome DNA exit site and partially detaching the DNA from the histone octamer. We use molecular dynamics simulations of the nucleosome to show that in this position the enzyme active site is perfectly poised to bind the H3 tail for deacetylation and that due to the unwrapping of the DNA even lysines close to the core of H3 can reach the enzyme. Furthermore, we find conformational changes, facilitated by the flexibility in SIRT6 between the zinc finger and the Rossmann fold, that could serve to fine-tune the position of the active site with respect to the H3 tail in order to trap its target acetylated residues.

Results

Overall structure

To investigate how nucleosomes stimulate the activity of SIRT6 we prepared nucleosome with the Widom-601 positioning sequence and reconstituted its complex with full-length recombinant human SIRT6 that was overexpressed in and purified from E. coli. The complex was stabilized via mild crosslinking before its deposition on cryo-EM grids.

We obtained a single particle cryogenic electron microscopy structure showing the nucleosome at 2.9 Å overall resolution and a protein density bound to its disc-like surface (Fig. 1a and Supplemental Fig. 2 and 3). SIRT6 could be fitted into this density with its zinc finger domain (resolved at 3.0 Å resolution), associated with the acidic patch of the nucleosome and its Rossmann fold domain (4.0-6.0 Å resolution) interacting with the DNA terminus at superhelix locations (SHL) 6 and 7 (Fig. 1b and Supplemental Fig. 4). The active site of SIRT6 is thus positioned just above where the H3 tail enters the nucleosome disc between the two DNA gyres. pseudosymmetric orientations of the Widom-601 DNA (Supplemental Fig. 5a). We find that SIRT6 is bound predominantly at the “right” TA-poor arm of the Widom-601 sequence where DNA interactions with the histone octamer are looser (Ngo et al., 2015). We detect however a small population with a second SIRT6 bound at the “left” TA-rich tight arm (Supplemental Fig. 2).

Structure of human SIRT6 in complex with the nucleosome

a, Front (top) and side (bottom) views of a composite cryo-EM reconstruction of human SIRT6-nucleosome. Maps from focused refinements of SIRT6 (magenta – Rossmann fold and Zn-finger domains labelled) and the nucleosome (H2A – yellow; H2B – orange; H3 – blue; H4 – green; DNA – light and dim grey). b, Corresponding views of the atomic model of the complex.

Conformational heterogeneity analyses in cryoDRGN (Zhong et al., 2021) reveal two principle large movements within the structure (Fig. 2a,c). First, we find that where the SIRT6 Rossmann fold binds, DNA is pulled away and detached from the histone octamer (Fig. 2b). Particles greatly vary in the extent of this shift that therefore probably represents a continuous movement of the DNA (compare Fig. 2b1 with 2b2). The Rossmann fold moves with and is attached to the DNA whereas the zinc finger domain remains fixed at the acidic patch of the nucleosome. In the full extent of the DNA movement, two turns from SHL 6 and 7 are detached from the histone octamer. This alters the path of the terminus DNA by 37° and breaks DNA interactions with histones H2A, H2B, and H3 (Supplemental Fig. 5b). Such a large change in DNA trajectory is reminiscent of the chromatin remodelling ATPases or pioneer transcription factors binding to nucleosome but was not observed in other histone modifiers (Dodonova et al., 2020; Farnung et al., 2017). Importantly, the DNA terminus without bound SIRT6 or free nucleosomes in our dataset did not show any deviation from the canonical conformations (Supplemental Fig. 5c). In addition, our MD simulations indicate higher flexibility of the DNA termini when SIRT6 is bound to the nucleosome (Supplemental Fig. 6). All together these findings strongly suggest that the DNA movement we observed is coupled to SIRT6 interactions. Second, in particles where the SIRT6-bound DNA remains in its canonical position, we observe shifts by up to 20 Å in the position of the Rossmann fold domain of SIRT6, partially loosening its association with the DNA (Fig. 2d). We show that this shift could be involved in fine tuning the position of the active site relative to the H3 tail.

Flexibility of the SIRT6 Rossmann fold

a, UMAP projection of the latent embeddings of a subset of SIRT6-nucleosome particles representing the concerted movement of the Rossmann fold and the DNA terminus. b, Structural representation of the two endpoints of the latent embeddings shown in panel a. c, Overlay of the DNA ends and the SIRT6 Rossmann fold of two endpoints shown in panel b. Arrows show their displacements between the two endpoints. d, UMAP projection of the latent embeddings of another subset of SIRT6-nucleosome particles representing the movement of the Rossmann fold with respect to the nucleosomal DNA. e, Structural representation of the two endpoints of the latent embeddings shown in panel b. f, Overlay of the DNA ends and the SIRT6 Rossmann fold of two endpoints shown in panel e. Arrow show the displacement of the Rossmann fold between the two endpoints.

These conformational changes are made possible by a flexible hinge region between the two domains of SIRT6, which is unique to this member of the sirtuins family.

Zinc finger domain binds the acidic patch

An important feature of the nucleosome surface is a cluster of acidic residues (E56, E61, E64, D90, E91, E92 of H2A and E102, E110 of H2B) that forms a negatively charged ‘acidic patch’ and takes the shape of a narrow groove. It serves as a docking site for many nucleosome-binding proteins. Four side chains of the SIRT6 zinc-finger domain extend into different sections of the acidic patch groove and interact with the negatively charged residues. R175 packs against H2B L103 and serves as a classical “arginine anchor motif” (McGinty and Tan, 2016), binding to the acidic triad of E61, D90 and E92 from H2A. SIRT6 presents two other arginines, R172, to contact E56 and Q24 of H2A as well as Q44 and E110 of H2B and R178 to contact H2A E91 and E102 from H2B. Finally, K170 interacts with E64 and N68 from H2A (Fig. 3a).

Binding of SIRT6 to the nucleosome

a, Close up view on zinc-finger interactions with the acidic patch. Color code as in Fig 1. b, Protein sequence alignment of human sirtuins. Red boxes and asterisks depict the residues of SIRT6 interacting with the acidic patch of the nucleosome. c, Protein sequence alignment of SIRT6 from different species highlighting the same amino acids as in panel b. Organisms: mm Mus musculus, rn Rattus norvegicus, bt Bos taurus, hs Homo sapiens and cc Castor canadensis. d, Depiction of the 3 arginines of SIRT6 (magenta) interacting with the DNA (grey). e, H2A c-terminal tail (yellow) interacts with SIRT6(magenta). f, SIRT6 binding to H2A or H2AZ containing nucleosomes. Bars show the fraction of residual nucleosomes that did not shift with bound SIRT6 in an electron-mobility shift assay (Supplemental Fig. 9). Bars represent mean ± SD of three biological replicates (shown as dots). ** indicates a statistically significant difference between the fraction of residual H2A and H2A.Z containing nucleosomes (P=0.0016 in paired t-test).

Substituting these four positively charged residues to alanines (SIRT6-4A) significantly weakened the interaction of SIRT6 with the nucleosomes as shown by electrophoretic gel-shift assay (Supplemental Fig. 7a). To test also the activity of SIRT6-4A we first acetylated the H3 tails of the nucleosomes with purified SAGA complex. The mutant protein showed strongly diminished activity in deacetylating H3K9Ac, further corroborating the role of the four basic residues R175, R172, R178 and K170, in binding to the nucleosome (Supplemental Fig. 7b).

All four SIRT6 positive residues are located in a loop that makes a sharp turn, due to a conserved glycine at position 173, connecting two beta strands and penetrates the acidic patch. Interestingly, this loop appears to be flexible in crystal structures of SIRT6 but rendered stable by its association with the acidic patch. Binding to different sections of the acidic patch maintains a strong hold of the zinc finger at the surface of the nucleosome and orients the Rossmann fold domain towards the DNA terminus. This strong hold is maintained throughout the movements of the Rossmann fold domain (Fig. 2).

It is important to note that the four positive residues of SIRT6 that bind the nucleosome acidic patch are conserved among vertebrates but in contrast are not present in other human sirtuins (Fig. 3b,c). This probably reflects the fact that SIRT6 is the only sirtuin predominantly associated with chromatin.

Rossmann fold domain binds nucleosome DNA exit site

Due to its flexibility with respect to the fixed zinc finger domain, the resolution of the Rossmann fold domain is not sufficiently high to distinguish most individual side chains. However secondary structure elements, mostly helices, are clearly visible and guided the precise fitting of the crystal structure of this domain.

We find that three highly conserved arginines from the Rossmann fold domain (R220, R231, R248) bind the phosphate backbone at SHL 6 and 7 of the nucleosomal DNA (Fig. 3d). These interactions are mostly maintained throughout the movement of the DNA terminus with the Rossmann fold domain bound. However, by comparison to dedicated DNA binding proteins such as transcription factors or TBP, the number of observed DNA-protein interactions is low in SIRT6. A biochemical study suggests that the intrinsically disordered C-terminal domain of SIRT6 is necessary for establishing tight interaction of SIRT6 with the nucleosome (Liu et al., 2020) via its ability to bind DNA. Our maps do show an additional isolated protein density that we could not assign with confidence to SIRT6 but associates with the DNA terminus Supplemental Fig. 8a). We propose however that the three arginines are required to correctly position the Rossmann fold, and hence the catalytic site, with respect to the H3 tail. Indeed, applying SAGA-acetylated nucleosomes as substrates, we find that a SIRT6 mutant, where the three arginines are substituted to alanines (SIRT6-3A), has severely diminished H3K9Ac deacetylase activity (Supplemental Fig. 7b). These findings confirm the importance of the three lysines R220, R231 and R248 to the enzymatic function.

Unlike other sirtuins, SIRT6 binds NAD+ in the absence of an acetylated substrate. This is attributed to the fact that crystal structures of SIRT6 describe a rigid helix (α3) where other sirtuins contain a flexible cofactor binding loop (Pan et al., 2011). In our map, we do not observe helix α3 (Supplemental Figure 8b) and we speculate that in the context of the nucleosome, this helix assumes a flexible loop conformation that like in other sirtuins will bind the cofactor only after stabilization by the acetyl lysine substrate.

In addition, we find that the N-terminal part of the Rossmann fold domain associates with the C-terminal tail of histone H2A at residue K119 (Fig. 3e). This observation suggests that monoubiquitination of H2A at this position, a major histone modification in mammalian cells, could interfere with the binding of SIRT6 or regulate its activity. Furthermore, we used electrophoretic mobility shift assay to follow the binding of SIRT6 to nucleosomes harboring the histone variant H2A.Z which deviates considerably from the canonical H2A in its C-terminal tail but not in other residues that contact SIRT6. Surprisingly, we find that binding of SIRT6 to nucleosomes is mildly (but statistically significant) stronger when H2A.Z is incorporated instead of H2A (Fig. 3f and Supplemental Fig. 9). This finding could be related to SIRT6’s role in repressing gene expression because H2A.Z is enriched in nucleosomes that flank the transcription start site.

SIRT6 is poised to deacetylate the H3 tail

The classical model of the nucleosome posits an elongated and disordered histone H3 N-terminal tail that freely roams and extends away from the core. However recent NMR studies and molecular dynamics predictions show that the tail actually collapses onto the nucleosome core, driven by robust interactions with DNA(Ikebe et al., 2016; Stutzer et al., 2016). The tail adopts multiple structurally heterogeneous, but energetically similar, states. It dynamically transitions between these conformations that all interact with the last superhelix location of nucleosomal DNA (Morrison et al., 2018). We used a published set of 15 predicted H3-tail conformations resulting from end-states of molecular dynamics simulations of the whole nucleosome (Morrison et al., 2018), to analyze whether the position of SIRT6 on the nucleosome serves the enzyme’s role in trapping residues of histone H3 tail. The nucleosome from each molecular dynamics simulation was superposed on our model of SIRT6 bound to a nucleosome with the DNA terminus at the canonical position. To represent the full range of movement of the Rossmann fold domain relative to the DNA we used two endpoints of this movement namely fully bound to DNA and partially dissociated from it (Fig. 4a and b, respectively). We then examined the position of the SIRT6 active site with respect to the predicted paths of the histone H3 tail.

SIRT6 poised to deacetylate lysine residues of H3

a, Side View of SIRT6 bound to the nucleosome with histone H3 K9 residue (red sphere) closest to the SIRT6 active site (blue spheres) from a set of molecular dynamics simulations. b, Side View of SIRT6 bound to the nucleosome with histone H3 K18 residue (orange sphere) closest to the SIRT6 active site from a set of molecular dynamics simulations. c, All H3K9 positions (red spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. d, All H3K18 positions (orange spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. e, Molecular dynamics simulations show that H3 c-terminal tail (blue) can protrude towards SIRT6 in a space formed between the histone octamer and the DNA. H3K27 is shown in red.

We find that in several H3 conformations, the tail is located in close proximity (<15Å) to the SIRT6 active site. In some of these cases, H3K9 or H3K18, the main target residues of SIRT6 (Wang et al., 2016), are practically situated in the active site. Furthermore, when analyzing the two endpoints of the Rossmann fold domain shift with respect to the DNA, we observe that in some H3 conformations, this shift draws the active site closer to the target residues (Fig. 4c,d). We conclude that the high-affinity binding of SIRT6 to the nucleosome poises the active site to trapping its main target residues. In addition, the flexibility of the Rossmann fold domain serves as a fine-tuning mechanism that allows for short-range scanning of the tail in order to locate the acetylated residues. We note also that except H3 no other histone is present at the vicinity of the SIRT6 active site. This gives rise to selectivity of SIRT6 towards the H3 tail.

To shed light on possible benefits offered by SIRT6-mediated DNA unwrapping we conducted molecular dynamic simulations of a nucleosome with bound SIRT6. Strikingly, new H3 tail conformations appear in the simulations that are absent from a control run with nucleosome alone. These pathways have the H3 tail protruding out of the histone core through the space between the DNA and the octamer that is formed via partial unwrapping of the DNA by SIRT6. Importantly, in these additional conformations H3 tail does not need to circle around the DNA to reach SIRT6 and therefore lysines that are closer to the core of H3 (e.g., K27) could also be accessible to the enzyme (Fig. 4e). To lend support to this claim we tested whether SIRT6 can deacetylate residue H3K27 that was first acetylated by SAGA (Supplemental Fig. 7c). We find that indeed SIRT6 could efficiently deacetylate H3K27Ac, although at a somewhat slower rate than H3K9Ac. We conclude that partial DNA unwrapping by SIRT6 allows H3-tail conformations that make lysines that are close to the core of H3 accessible to the enzyme.

Discussion

Most histone-modifying enzymes, including sirtuin family members such as SIRT1, show significant activity on isolated target peptides and form multiple interactions with the residue to be modified as well as with flanking amino acids to provide selectivity. It is therefore striking that the measured turnover rate for SIRT6 on isolated H3K9 peptides is 1000 slower than for other sirtuins. Moreover, unlike SIRT1, the activity of SIRT6 is dramatically stimulated by binding to its cognate nucleosome substrate.

We solved the structure of SIRT6 bound to a nucleosome and used heterogeneity analysis to uncover the main conformational changes in the complex. Examining these findings in the context of molecular dynamics studies of the nucleosome that produce putative paths of histone tails, shed light on the unique properties of SIRT6.

We find that multiple interactions of SIRT6 with the nucleosome position its active site at the putative path of relatively long-lived conformations of H3 so that residues K9 and K18 of the tail spend significant time very close to, or even at, the enzyme active site. In addition, SIRT6 moves with respect to the nucleosomal DNA allowing it to fine-tune the position of the active site to trap the target residue. Hence it is the binding of SIRT6 to the nucleosome and an intrinsic scanning capability that account for its much higher enzymatic efficiency on nucleosomes. Importantly, the position of SIRT6 also ensures high selectivity because the space surrounding the active site is devoid of any other tails apart from H3.

The recognition mode of nucleosomal H3 employed by SIRT6 is reminiscent of that used by the PRC1 complex in the specific ubiquitylation of H2A Lys119 (McGinty et al., 2014). Similar to SIRT6, PRC1 requires a nucleosome substrate, showing no activity on an H2A-H2B dimer, and multiple interactions with the nucleosome position an active site in close proximity to its target residue. However, the flexibility and scanning capacity we observed for SIRT6 is lacking in PRC1 perhaps echoing the fact that its target residue is rather stable in the nucleosome compared to the H3 tail.

It is interesting to note that the flexibility of SIRT6 that is at the root of its scanning mechanism was suggested to be, at least in part, responsible for its nearly undetected activity on isolated peptides (Pan et al., 2011). It is only when it is constrained to the proximity of the target residues, through binding to the nucleosome, that this capacity facilitates the function of the enzyme.

A second major conformational change observed in our 3D reconstructions is the detachment of the DNA terminus, with a bound Rossmann fold domain, from the octamer. Our molecular dynamics simulations suggest that the space formed between DNA and octamer allows for additional conformations of the H3 tail that protrudes towards SIRT6. These new pathways do not require circling around the DNA gyre and therefore lysines that are closer to the core of H3 (e.g., K27) can reach in this way the active site of SIRT6. We note that K56 which lies at the core of H3 within a stable helix was also reported to be a target for SIRT6. It is therefore possible that the enzyme employs additional recognition modes of its targets.

Because the position of the SIRT6 active site seems to be well suited for modifying the H3 tail, it stands to reason that other H3 modifying enzymes, such as the GCN5 HAT module of SAGA and ATAC, would place their active site in a position similar to the one of SIRT6 and possibly also displace the DNA from the histone octamer. Gcn5 deviates from SIRT6 in having significant affinity for the H3 tail and therefore might pose a less stringent requirement for precise positioning of the active site or for specific conformations of the H3 tail. Thus, we speculate the Gcn5 recognition mode will be similar to SIRT6 but will also include some sequence specific binding. While this manuscript was under review two studies describing the structure of SIRT6-nucleosome appeared in press (Chio et al., 2023; Wang et al., 2023). The conclusion of these papers regarding the position of SIRT6 on the nucleosome and the unwinding of DNA by the enzyme are similar to our findings. We however dissected in addition the movements of SIRT6 on the nucleosome and analyzed via molecular dynamics the conformations of the H3 tail with respect to the SIRT6 active site. Our results point to the importance of the flexibility between the globular domains of SIRT6 and also explain how SIRT6 can access lysines that are much closer to the histone core than H3K9.

Methods

Human SIRT6 purification

Plasmid (AddGene #28271) was transformed into E. coli strain BL21-RIL. Cells in 1L TB (with 50 μg/ml Kanamycin and 25 μg/ml Chloramphenicol) grew at 37C degrees till OD260 reached 0.8 when they were rapidly cooled on ice to 18C. Over-expression was induced by the addition 0.5 mM IPTG and cells continued to grow at 18C over-night. All subsequent steps were performed at 0-4C and solutions were always supplemented with a protease inhibitor cocktail (PMSF, Peptstatin, E-64). Cells were harvested by centrifugation at 8250g for 20 minutes, suspended in buffer A200 (20 mM Hepes pH 7.5, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 200 mM NaCl) and treated with lysosyme (1 mg/ml) for 30 minutes on ice. Salt concentration was increased to 0.5M and cells were then lysed by sonication (6*10sec). Debris were pelleted by centrifugation 45,000g*20minutes. The supernatant was incubated with 0.5 ml of Ni beads (1 ml slurry, Roche), equilibrated in buffer A500, for 1h. The solution was poured into a gravity column (CliniScience). Beads were then washed by 10 ml of buffer A500 with 5mM imidazole before elution by 2ml of buffer B300 (20 mM Hepes pH 7.2, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 300mM NaCl) with 150mM imidazole. The eluted protein was diluted with two volumes of buffer D (20 mM Hepes pH 7.0, 5% glycerol v/v, 2 mM beta-mercaptoethanol), loaded on a 1ml Heparin column and eluted with 0.1M-1M salt gradient in buffer D. Fraction were analyzed by SDS-PAGE and those containing SIRT6 were pooled and dialyzed against buffer E (20 mM Hepes pH 7.5, 2% glycerol v/v, 0.5 mM TCEP and 150 mM NaCl. Human SIRT6 was then concentrated to 7.8 mg/ml, flash frozen in N2, and kept in aliquotes at -80C.

Cloning of SIRT6 mutants

SIRT6 mutants were created as follows: the region of AddGene plasmid #28271 between SacI and HindIII sites were synthetized by SynBio company with changes in DNA sequence leading to K170A, R172A, R175A, R178A, R220A, R231A and R248A mutations in amino acid sequence. A NheI restriction site was also introduced between the first four and last three mutations without changes in amino acid sequence in order to facilitate cloning. A plasmid expressing SIRT6-7A that incoporates all seven mutations was created via ligation of the synthetic DNA fragment into the original expression plasmid between the SacI and HindIII sites. Fragments of the original expression plasmid corresponding to the regions between HindIII and NheI or NheI and SacI were amplified with PCR and were cloned separately through corresponding restriction sites into the plasmid expressing SIRT6-7A to yield plasmids expressing SIRT6 with the first four mutations (SIRT6-4A) or the last three mutations (SIRT6-3A), respectively.

Octamer reconstitution and nucleosome DNA

Octamers were reconstituted from individual Xenopus leavis (canonical) or human (H2A.Z) histones expressed as inclusion bodies according to the standard protocol (Dyer et al., 2004; Luger et al., 1999). Widom 601 145bp DNA was produced using a plasmid harboring 16 copies of this sequence as described by Dyer et al. (Dyer et al., 2004).

Nucleosome core particle reconstitution

Nucleosomes with 145bp Widom 601 positioning sequence were prepared according to NEB Dilution Assembly Protocol (E5350)(https://international.neb.com/protocols/2012/06/02/dilution-assembly-protocol-e5350) with some modifications as follows: 2.75 μM 145 bp Widom601 DNA was mixed with 2.75 μM canonical or H2AZ histone octamers in a solution containing 2M NaCl, 1mM EDTA, 5mM beta-mercaptoethanol. The solution was incubated for 30 min at RT and then underwent serial dilutions down to 1.48M, 1M, 0.6M, 0.25M NaCl with buffer LowSalt (10 mM Hepes-KOH pH 8.0, 2.5 mM beta-mercaptoethanol). After each dilution the solution was kept at RT for 30 min. In order to reduce the final NaCl concentration, nucleosomes were concentrated in 0.5 mL 100KDa cutoff Amicon up to 100 μl, then diluted 5 times with buffer LowSalt. This step was repeated one more time. Finally, nucleosomes were concentrated to 3-4 μM and analyzed in a 5% native 0.2xTBE polyacrylamide gel to ascertain the quality of the sample and absence of free DNA.

Nucleosome acetylation

Canonical W601 nucleosomes (final concentration 1.8 μM) were mixed with the holo-SAGA complex (final 18 nM) purified from Pichia pastoris cells according to (Papai et al., 2020) in 20 mM Hepes-KOH pH 7.5, 10% v/v glycerol, 50 mM NaCl, 2 mM MgCl2, 5 mM beta-mercaptoethanol, 0.005% Tween-20, and 0.05 mM Acetyl-CoA. Reaction was incubated on ice for 16 hours. On the next day SAGA was removed via incubation of the mixture with Streptavidin Sepharose beads. In order to reduce the remaining Acetyl-CoA concentration, nucleosomes were concentrated in 0.5 mL 100KDa cutoff Amicon up to 100 μl, then diluted 5 times with buffer 10 mM Hepes-KOH pH 8.0, 2.5 mM beta-mercaptoethanol. This step was repeated four more time. Finally, the nucleosomes were concentrated to 3.5 μM.

Binding and gel-shift assays

For nucleosome-binding assays, 100 nM of recombinant nucleosomes (harboring either H2A or H2AZ histone) were incubated in binding buffer (10 mM Hepes, 50 mM NaCl, 0.5 mM TCEP, 6% glycerol, and pH 7.5) with SIRT6 at 1:2 molar ratio. The reactions were performed on ice for 1h in LoBind tubes (Eppendorf) to minimize protein loss. The reactions were then resolved in a 0.2x TBE 5% gel. The gel was stained with Ethidium Bromide. Bands were quantified with ImageJ.

Deacetylation assay

Non-acetylated or SAGA-acetylated canonical W601 nucleosomes were mixed with SIRT6 (wt, SIRT6-4A or SIRT6-3A) at 100 nM and 40 nM, respectively, in 50 mM Hepes-KOH pH 7.5, 1 mM DTT, 0.2 mg/mL BSA, 1 mM NAD+. Reactions were set on ice and then incubated at 37 °C for 1 or 2 hours. Reactions were quenched by boiling with 5X Laemmli buffer. The proteins were separated on 15% SDS-PAAG and acetylation was analyzed by Western blot using the following antibodies: For H3K27Ac CellSignaling #8173S diluted 1:1000; For H3K9Ac Abcam Ab4441 1:2500; For H3 CellSignaling #14269S 1:1000.

Grid preparation

SIRT6 was first diluted in buffer (20mM Hepes-K pH 7.5, 150mM NaCl, 0.5 mM TCEP, 2% Trehalose). Nucleosome and SIRT6 were then mixed at a 1:2 molar ratio. Final conditions included 1.7 μM Nucleosome and 3.4 μM SIRT6 in buffer (20mM Hepes-K pH 7.5, 18mM NaCl, 0.5mM TCEP). After 1h on ice glutaraldehyde and the detergent dodecyl-maltoside were added at a final concentration of 0.05% v/v and 0.0125% w/v, respectively. The sample was incubated with the cross-linker for an additional 30-60 minutes. 2nm carbon foil was floated on 3.5/1 300 mesh Quantifoil grids. 3 μl of sample was deposited onto such grids, blotted 5s with blot force 5 after 60s incubation in a Vitrobot IV (Thermo Fisher) at 6 °C using 100 % humidity and flash-frozen in liquid ethane.

cryo-EM data acquisition

Two different datasets were collected. For the first dataset images were acquired on a Cs-corrected Titan Krios G1 (Thermo Fisher) microscope operating at 300 kV in nanoprobe mode using serialEM for automated data collection. Movie frames were recorded on a Gatan K3 direct electron detector after a Quantum LS 967 energy filter using a 20 e-V slit width in zero-loss mode. Images were acquired hardware-binned at a nominal magnification of 130,000, which yielded a pixel size of 0.862 Å. Forty movie frames were recorded at a dose of 1.12 and 1.30 e− per Å2 per frame. The second image dataset was acquired on a Titan Krios G4 microscope operating at 300 kV in nanoprobe mode using EPU for automated data collection. Movie frames were recorded on a Flacon 4i direct electron detector after a Selectris X energy filter using a 10 e-V slit width in zero-loss mode. Images were acquired at nominal magnification of 270,000, which yielded a pixel size of 0.458 Å.

Image analysis

For the first dataset movie frames were aligned, dose-weighted in Warp (Tegunov and Cramer, 2019) together with CTF estimation and particles picking. For the second dataset frame alignment, dose-weighting and CTF estimation was done in cryoSPARC (Punjani et al., 2017), picking was done with crYOLO (Wagner et al., 2019). These datasets were analyzed in cryoSPARC and RELION 3.1 (Zivanov et al., 2018) according to standard protocols. Briefly, images were subjected to reference-free 2D classification in cryoSPARC to remove images with contaminations. Ab-initio reconstruction was performed on the selected images followed by heterogenous and homogenous refinements in cryoSPARC. Structural variability and flexibility were explored with Relion 3.1 using 3D classification without alignment with different T values and cryoDRGN (Zhong et al., 2021) according to standard protocols. Sharpening of the maps was done with DeepEmhancer (Sanchez-Garcia et al., 2021).

Model building and refinement

A composite map was created from the part corresponding to the SIRT6 Rossmann-fold from structure of the first dataset and the SIRT6 Zn-finger plus nucleosome from the structure of the second dataset. Crystal structure of Widom 601 sequence containing nucleosome (PDB: 3LZ0) and human SIRT6 (PDB: 5×16) were rigid-body fitted into this composite cryo-EM structure in Chimera (Pettersen et al., 2004). Flexible-fittings were performed with Isolde (Croll, 2018) in ChimeraX (Goddard et al., 2018) into the cryoDRGN structures. All models were refined in real-space with PHENIX (Liebschner et al., 2019) and Isolde.

Molecular Dynamics simulations

All molecular dynamics (MD) simulations were performed with the AMBER20 suit of programs, using the ff14SB (Maier et al., 2015) force field with bsc1 DNA corrections (Ivani et al., 2016) and CUFIX ions corrections (Yoo and Aksimentiev, 2018) that improve the description of disordered region conformations. Parameters for the zinc finger were taken from Macchiagodena et al. (Macchiagodena et al., 2020).

The histone tails were added to the cryo-EM structure based on the work described by Armeev et al (Armeev et al., 2021), and protonation states of the protein residues were checked with the propka3 software (Olsson et al., 2011). The system was soaked in a truncated octrahedral TIP3P water box of 20Å buffer and 0.15 mM of NaCl ions were added to simulate physiological conditions, resulting in a total of 330,000 atoms.

The minimization of the energy was carried out in 5 steps, each one alternated with a 20ps NVT equilibration step at 100K to ensure a very smooth optimization of the starting geometry. Each minimization run included 5,000 steps in steepest descent and 5,000 steps in conjugate gradient, with decreasing restraints on the protein and nucleic atoms from 20kcal/mol to 5kcal/mol that were kept in the associated equilibration step. The last minimization step was released of any constraints and followed by a final NVT equilibration run at 300K. The time step was then switched from 2 to 4fs as allowed by the use of the SHAKE bond length constraints (Kräutler et al., 2001) and the Hydrogen Mass Repartioning algorithm (Hopkins et al., 2015). The latter was applied on the topology using the parmed tool of Amber20. A 40ns equilibration run in NTP was then carried out, followed by a 5 μs production. Coordinates were outputted every 1ns during the production run. The temperature was kept constant using the Langevin thermostat with a 2ps-1 collision frequency, and the Berendsen barostat was used to maintain the pressure at 1 atm in the NTP runs. Electrostatic interactions were treated using the Particle Mesh Ewald approach (Darden et al., 1993) with an 8Å cut-off. Three replicates were performed starting with different velocities at the unconstrained 300K NVT equilibration step, resulting in a total of 15 μs of sampling.

A cluster analysis was performed on the MD ensembles based on the DNA and protein RMSD. Conformations were extracted every 5 frames of the trajectories to avoid memory overload and were sorted using the hierarchical agglomerative and average linkage algorithms. 10 clusters were generated by replicate (Supplemental Table 2).

The DNA flexibility analysis was performed on each MD ensemble by using a Principal Component Analysis-based machine learning script quantifying the contribution of each residue to the overall dynamics of the system. Based on the Scikit-learn58 library (Fleetwood et al., 2020), it calculates a covariance matrix from the internal coordinates of the MD ensemble. In this matrix, the eigenmodes showing the highest amplitude are the most implicated in the main fluctuations of the system, and the contribution of each residue is normalized based on these values.

The control system was generated by removing SIRT6 from the initial structure and was submitted to the same protocol.

Data availability

The cryo-EM maps have been deposited in the Electron Microscopy Data Bank (EMDB) under accession codes EMD-16842 (High-resolution on the nucleosome), EMD-16843 (SIRT6-nucleosome refined for SIRT6) and EMD-16845 (composite map). The model coordinates for SIRT6-nucleosome were deposited in the Protein Data Bank (PDB) under the accession code 8OF4.

All MD input files are publicly available on Github: https://github.com/emmanuellebignon/NCP-SIRT6

Supporting information

eLife Assessment

This manuscript provides a useful reconstruction of the structure of the sirtuin-class histone deacetylase Sirt6 bound to a nucleosome based on cryo-EM observations, and additional characterization of the flexibility of the histone tails in the complex based on molecular dynamics simulations. Similar structures have recently been published, but this work provides solid support for the conclusions of those papers and also includes some novel insights into the potential dynamics of Sirt6 bound to a nucleosome that help explain its substrate specificity.

Joint Public Review:

Smirnova et al. present a cryo-EM structure of human SIRT6 bound to a nucleosome as well as the results from molecular dynamics simulations. The results show that the combined conformational flexibilities of SIRT6 and the N-terminal tail of histone H3 limit the residues with access to the active site, partially explaining the substrate specificity of this sirtuin-class histone deacetylase. Two other groups have recently published cryo-EM structures of SIRT6:nucleosome complexes; this manuscript confirms and complements these previous findings, with the addition of some novel insights into the role of structural flexibility in substrate selection.

This manuscript is the third in a recent series of reports of cryo-EM structures of Sirt6:nucleosome complexes. The main conclusions of the three studies are similar, but this manuscript from Smirnova et al. includes additional molecular dynamics analysis of the histone tails. These studies suggest that part of the specificity for sites on the H3 tail is the result of only this tail having significant access to the active site. The results are partially validated by showing that H3-K27Ac is sometimes found near the active site in the simulations, and is a weak substrate for the deacetylase in vitro. All of the structures show Sirt6 contacting the acidic patch of H2A-H2B, partial displacement of the H2A C-terminal tail, and displacement of the DNA at the entry-exit site to "unclamp" the H3 N-terminal tail. This manuscript provides additional support for the conclusions drawn in the first two published structures, adds molecular dynamics simulations that provide further insight and includes a biochemical assay that helps to resolve an apparent conflict regarding the deacetylation of H3-K27Ac from the other two papers.

Author Response

The following is the authors’ response to the original reviews.

We are grateful to the reviewers for their remarks, which significantly improved the paper. We repeated the biochemical assay concerning SIRT6 activity on H3-K27Ac and quantified the results as requested. Please find our detailed answers bellow each recommendation of the reviewers.

Major recommendations:

1. Grammatical errors are still common; the authors may need to consider an external editing service if they intend to fix the problems as they indicate that they believe the errors have been removed. The Results section is relatively clean, but parts of the Abstract, Introduction, and Discussion are more difficult to understand, and errors are especially common in the Methods section and those parts of the manuscript that are new in this revision.

We corrected the grammatical errors.

1. The introduction doesn't mention the other structures published; this is considered to be a serious deficiency as it prevents the reader from understanding the context for the contributions described here. Withholding the comparison with (or mention of) the previously published work to the last sentence of the Discussion seems misleading and does not give the reader adequate ability to judge the novelty of the results presented in this manuscript.

A paragraph comparing our paper to the other structures published appear at the end of the discussion. We feel this is still the right place for such a paragraph.

1. The addition of the assay for deacetylation is a significant improvement over the initial submission. This is important both for validating the importance of the acidic patch contacts and for helping to resolve the conflicting reports regarding activity on H3-K27Ac. Given the importance of this assay for the impact of the manuscript, it is not clear why the authors chose to 1) put the data in the supplement instead of in the main manuscript, and 2) provide only single samples without quantitation. These both seem to be significant limitations.

We repeated the experiment and provided quantification of the results. We placed the figure in the main manuscript.

1. The authors should add text or a table to the Methods section explaining which maps were used for each figure. By our count, there are 8 maps and 5 models (plus MD models) based on two datasets, but the relationships among them are not clearly stated, and the names of the maps (such as "Zn-finger focused" and "Rossman-Fold-Focused") might be changed to be more helpful to the reader (for example, the latter includes more than the Rossman fold and might be renamed "Sirt6-focused"). The authors should also explain how the maps were validated, which data were deposited in public repositories, and why some data were not deposited. For example, no statistics or methods regarding how particles were separated into integrated vs. non-integrated motion are provided for the CryoDRGN models. Further, the "two principle movements" described are depicted in 4 maps from two CryoDRGN runs using two separate sets of particles, but the relationships among them are not defined clearly. Finally, the connectivity of densities in Fig 8 are not obvious in the submitted maps. Until these points are addressed, the work is considered incomplete.

AND

1. The PDB model provided for review and submitted to the PDB database shows loosely bound DNA at the nucleosomal entry/exit points near the binding site of SIRT6, but the maps provided for review and submitted to the EMDB show stronger density for the canonical location of the DNA expected at these sites. The CryoDRGN maps support a more extended conformation, but these maps were not deposited or provided for review so their validity cannot be assessed.

We added a section to the methods listing the different maps used for the figures. We deposited the map we used to trance the H2A N-terminal tail (EMD-18497). Unfortunately, we couldn’t deposit the cryoDRGN maps as the deposition system either accepts composite maps, where the consensus should be deposited too or experimental maps, where the deposition of half maps are mandatory. Nevertheless, the cryoDRGN maps are available upon request. We also added a supplementary figure (Supplementary Fig 6) to show how the cryoDRGN analyses were performed.

1. The orientation, angle and threshold used in Fig 1 make it difficult to see the multiple DNA orientations that are visible in the deposited consensus map. Examination of the map suggests that the DNA model submitted to PDB corresponds to a weaker DNA conformation than is present in the map where both DNA conformations are visible. The authors should consider modeling both conformations in their deposited model to provide a more complete, accurate representation of the data. It is concerning that a key conclusion of the manuscript is that the DNA conformation changes upon SIRT6 binding, but density for the canonical position is observable in Fig 8a.

Figure 1 is showing the overall representation of the SIRT6 bound nucleosome structure. We show the DNA linker orientations in the subsequent figure. Figure 8 (now Figure 9) shows the rearrangement of the SIRT6 Rossmann fold domain not the DNA linker.

1. Figure 4 needs a more complete legend, indicating that it is a hybrid of the consensus structure (one color) and the MD simulations (another color). In general, the colors used in the figure should be changed to make the main points more accessible.

As there is a color code for the histones, changing colors might be confusing. The figure legend mentions that panels c, d and e are from MD simulations.

Minor recommendations:

1. Figures 2c, e, and f are not referenced in the text.

We now referenced all figure panels in the text.

1. Consider moving Supp. 5C to Fig. 2 as the models in that figure come from the CryoDRGN maps and not the consensus map.

Supplemental Figure 5c show the DNA linker deviation upon SIRT6 binding from another angle. We prefer to keep it there.

1.) Supp Fig 3 is labeled "ZnF-nucleosome" refinement, but this appears to come from Data Set #2 processing. The map might be labeled ZnF-nucleosome but then a mask should be shown that excludes the Rossman Fold. It is not clear if this is a focused refinement or just a 2.9 A map that was merged with the "Rossman-fold" map.

We changed both supplemental figures accordingly.

1. The orientation of Fig 2 b and e do not show the differences in these models as well as panels c and f. Panels b and e could be replaced with the 4 CryoDRGN maps.

The models reflect the cryoDRGN maps and panels c and f were added to clarify the movement.

1. The MD description should emphasize that the H3 tails are moving with respect to the active site, as it currently suggests the active site is moving.

In the results and in the discussion section we mention that we observe new conformations of the H3 tail, not of the active site.

1. The authors refer to the "flexibility of the Rossmann fold domain," but the Rossman Fold domain isn't flexible, the linkage to the ZnF is flexible. Perhaps "observed conformational space" or "dynamic Rossman-fold domain position" are meant.

The text was changed accordingly.

1. The H2A C-terminal tail present in Fig 1 (bottom right) and Figure 3e is not present in the model in Fig 4a,b.

The H2A tails conformation was not resolved in the cryoDRGN maps so we didn’t model it.

1. The crosslinking agent used is not specified.

The crosslinking agent used is specified more clearly in the methods.

1. Supp Table 1 and EM methods do not agree on the magnification for Dataset #1. Verify nominal versus binned magnification and reported pixel size.

1. Fig 3F showing the difference between affinity for H2A and H2A.Z-containing nucleosomes would be more convincing with a titration rather than the current comparison of a single concentration.

We agree with this remark however, we find single concentration comparison is convincing enough for the purposes of this paper as it is not a central finding.

1. Fig S1 legend; both the Zn-finger and helix bundle are stated to be shown in green.

Figure S1 legend was changed.

Version 3

Abstract

Abstract

Sirtuin 6 (SIRT6) is an NAD+ dependent histone H3 deacetylase that is prominently found associated with chromatin, attenuates transcriptionally active promoters and regulates DNA repair, metabolic homeostasis and lifespan. Unlike other sirtuins, it has low affinity to free histone tails but demonstrates strong binding to nucleosomes. It is poorly understood how SIRT6 docking on nucleosomes stimulates its histone deacetylation activity. Here we present the structure of human SIRT6 bound to a nucleosome determined by cryogenic electron microscopy. The zinc finger domain of SIRT6 associates tightly with the acidic patch of the nucleosome through multiple arginine anchors. The Rossmann fold domain binds to the terminus of the looser DNA half of the nucleosome, detaching two turns of the DNA from the histone octamer and placing the NAD+ binding pocket close to the DNA exit site. This domain shows flexibility with respect to the fixed zinc finger and moves with, but also relative to, the unwrapped DNA terminus. We apply molecular dynamics simulations of the histone tails in the nucleosome to show that in this mode of interaction, the active site of SIRT6 is perfectly poised to catalyze deacetylation of the H3 histone tail and that the partial unwrapping of the DNA allows even lysines close to the H3 core to reach the enzyme.

Introduction

The state of lysine acetylation in histones, as well as in several other proteins, is determined by the balancing act of acetyltransferases and deacetylases. Sirtuin is an evolutionarily conserved family of lysine deacetylases that comprises seven members in humans and plays an important role in regulating numerous physiological and pathological processes including aging, cancer, inflammation and metabolism (Cen et al., 2011; Vitiello et al., 2017; Watroba et al., 2017). Unlike other lysine deacetylases sirtuins carry out their enzymatic reaction by transferring the acetyl group to an essential NAD+ cofactor before acetylating the target protein (Imai et al., 2000).

Human Sirtuin 6 (SIRT6) is a multitasking enzyme implicated in diverse pathways. Through its histone H3 deacetylation at gene promoters, it is a transcription repressor attenuating inter-alia ribosome biogenesis (Sebastian et al., 2012), glycolysis (Taylor et al., 2022) and the expression of various transcription factors, including nuclear factor κB (NF-κB) (Kawahara et al., 2009) and hypoxia-inducible factor-1α (HIF-1α) (Zhong et al., 2010). Its H3K9 deacetylation of telomeric chromatin maintains telomere integrity (Michishita et al., 2008). In addition, via its second enzymatic activity, ADP-ribosylation of PARP1 (Mao et al., 2011) and by stabilizing DNA-PK on chromatin (McCord et al., 2009), SIRT6 also fosters DNA double-strand break repair.

These activities position SIRT6 as a key regulator of longevity as indeed shown by several lines of evidence. In mice, knockout of SIRT6 leads to a shortened lifespan and a premature aging-like phenotype (Kanfi et al., 2012). Similar observations were made in monkeys deficient for the enzyme (Zhang et al., 2018). Conversely, overexpression of SIRT6 extends lifespan and diminishes frailty in mice (Roichman et al., 2021). Furthermore, a study with several rodent species found a positive correlation between SIRT6 ability to promote DNA double-strand break repair and longevity - in other words, more active SIRT6 leads to longer lifespan (Tian et al., 2019).

The development of cancer can also be regulated by SIRT6. However, whereas in some malignancies SIRT6 plays a role as a tumor suppressor (Marquardt et al., 2013; Min et al., 2012; Sebastian et al., 2012), in others it acts as an oncogene (Bauer et al., 2012; Khongkow et al., 2013; Liu et al., 2013). The molecular mechanisms underlying SIRT6 effects on longevity or on the two contradictory roles in cancer development are not well understood, limiting the search for medically useful drugs that target the enzyme.

SIRT6 is a 37 KDa protein composed of two globular parts, a Rossmann fold domain that harbors the NAD+ binding site and a zinc finger domain (Pan et al., 2011). A striking difference between SIRT6 and other members of the sirtuin family is the presence of a splayed zinc finger and the lack of a stabilizing helix bundle linking the two globular domains. This absence probably endows SIRT6 with a unique flexibility between the domains (Supplemental Fig. 1).

These structural differences are manifested in function. Unlike other sirtuins, SIRT6 has very low activity on acetylated free H3 histones or histone peptides (Jiang et al., 2013). This seemed to contradict findings concerning the function of SIRT6, in particular data demonstrating its activity in vivo as H3 deacetylase (Michishita et al., 2008) as well as the fact that SIRT6 is nuclear and found mainly associated with chromatin (Michishita et al., 2005). The conundrum was settled when it was revealed that nucleosomes, the in vivo cognate substrate of SIRT6, dramatically enhance the enzyme activity towards acetylated H3 histones (Gil et al., 2013). Further biochemical analysis showed that SIRT6 binds nucleosomes with high affinity via multiple interacting sites (Liu et al., 2020).

However, a mechanistic understanding of SIRT6 binding to nucleosomes and how this association promotes higher enzyme efficacy is currently lacking.

Here we present the structure of human SIRT6 bound to the nucleosome. We identify interactions between the zinc finger domain and the acidic patch of the nucleosome as well as between the Rossmann fold and the DNA terminus, positioning the NAD+ binding pocket close to the nucleosome DNA exit site and partially detaching the DNA from the histone octamer. We use molecular dynamics simulations of the nucleosome to show that in this position the enzyme active site is perfectly poised to bind the H3 tail for deacetylation and that due to the unwrapping of the DNA even lysines close to the core of H3 can reach the enzyme. Furthermore, we find conformational changes, facilitated by the flexibility in SIRT6 between the zinc finger and the Rossmann fold, that could serve to fine-tune the position of the active site with respect to the H3 tail in order to trap its target acetylated residues.

Results

Overall structure

To investigate how nucleosomes stimulate the activity of SIRT6 we prepared nucleosome with the Widom-601 positioning sequence and reconstituted its complex with full-length recombinant human SIRT6 that was overexpressed in and purified from E. coli. The complex was stabilized via mild crosslinking before its deposition on cryo-EM grids.

We obtained a single particle cryogenic electron microscopy structure showing the nucleosome at 2.9 Å overall resolution and a protein density bound to its disc-like surface (Fig. 1a and Supplemental Fig. 2 and 3). SIRT6 could be fitted into this density with its zinc finger domain (resolved at 3.0 Å resolution), associated with the acidic patch of the nucleosome and its Rossmann fold domain (4.0–6.0 Å resolution) interacting with the DNA terminus at superhelix locations (SHL) 6 and 7 (Fig. 1b and Supplemental Fig. 4). The active site of SIRT6 is thus positioned just above where the H3 tail enters the nucleosome disc between the two DNA gyres. The highresolution at the nucleosome allows discerning purines from pyrimidines. We could therefore distinguish between the two pseudosymmetric orientations of the Widom-601 DNA (Supplemental Fig. 5a). We find that SIRT6 is bound predominantly at the “right” TA-poor arm of the Widom-601 sequence where DNA interactions with the histone octamer are looser (Ngo et al., 2015). We detect however a small population with a second SIRT6 bound at the “left” TA-rich tight arm (Supplemental Fig. 2).

Structure of human SIRT6 in complex with the nucleosome

a, Front (top) and side (bottom) views of a composite cryo-EM reconstruction of human SIRT6-nucleosome. Maps from focused refinements of SIRT6 (magenta – Rossmann fold and Zn-finger domains labelled) and the nucleosome (H2A – yellow; H2B – orange; H3 – blue; H4 – green; DNA– light and dim grey). b, Corresponding views of the atomic model of the complex.

Conformational heterogeneity analyses in cryoDRGN (Zhong et al., 2021) reveal two principle large movements within the structure (Fig. 2a,d and Supplemental Fig. 6).

Flexibility of the SIRT6 Rossmann fold

a, UMAP projection of the latent embeddings of a subset of SIRT6-nucleosome particles representing the concerted movement of the Rossmann fold and the DNA terminus. b, Structural representation of the two endpoints of the latent embeddings shown in panel a. c, Overlay of the DNA ends and the SIRT6 Rossmann fold of two endpoints shown in panel b. Arrows show their displacements between the two endpoints. d, UMAP projection of the latent embeddings of another subset of SIRT6-nucleosome particles representing the movement of the Rossmann fold with respect to the nucleosomal DNA. e, Structural representation of the two endpoints of the latent embeddings shown in panel b. f, Overlay of the DNA ends and the SIRT6 Rossmann fold of two endpoints shown in panel e. Arrow show the displacement of the Rossmann fold between the two endpoints.

First, we find that where the SIRT6 Rossmann fold binds, DNA is pulled away and detached from the histone octamer (Fig. 2b). Particles greatly vary in the extent of this shift that therefore probably represents a continuous movement of the DNA (compare Fig. 2b1 with 2b2). The Rossmann fold moves with and is attached to the DNA whereas the zinc finger domain remains fixed at the acidic patch of the nucleosome (Fig. 2c). In the full extent of the DNA movement, two turns from SHL 6 and 7 are detached from the histone octamer. This alters the path of the terminus DNA by 37° and breaks DNA interactions with histones H2A, H2B, and H3 (Supplemental Fig. 5b). Such a large change in DNA trajectory is reminiscent of the chromatin remodelling ATPases or pioneer transcription factors binding to nucleosome but was not observed in other histone modifiers (Dodonova et al., 2020; Farnung et al., 2017). Importantly, the DNA terminus without bound SIRT6 or free nucleosomes in our dataset did not show any deviation from the canonical conformations (Supplemental Fig. 5c). In addition, our MD simulations indicate higher flexibility of the DNA termini when SIRT6 is bound to the nucleosome (Supplemental Fig. 7). All together these findings strongly suggest that the DNA movement we observed is coupled to SIRT6 interactions.

Second, in particles where the SIRT6-bound DNA remains in its canonical position, we observe shifts by up to 20 Å in the position of the Rossmann fold domain of SIRT6, partially loosening its association with the DNA (Fig. 2e,f). We show that this shift could be involved in fine tuning the position of the active site relative to the H3 tail.

These conformational changes are made possible by a flexible hinge region between the two domains of SIRT6, which is unique to this member of the sirtuins family.

Zinc finger domain binds the acidic patch

An important feature of the nucleosome surface is a cluster of acidic residues (E56, E61, E64, D90, E91, E92 of H2A and E102, E110 of H2B) that forms a negatively charged ‘acidic patch’ and takes the shape of a narrow groove. It serves as a docking site for many nucleosome-binding proteins. Four side chains of the SIRT6 zinc-finger domain extend into different sections of the acidic patch groove and interact with the negatively charged residues. R175 packs against H2B L103 and serves as a classical “arginine anchor motif” (McGinty and Tan, 2016), binding to the acidic triad of E61, D90 and E92 from H2A. SIRT6 presents two other arginines, R172, to contact E56 and Q24 of H2A as well as Q44 and E110 of H2B and R178 to contact H2A E91 and E102 from H2B. Finally, K170 interacts with E64 and N68 from H2A (Fig. 3a).

Binding of SIRT6 to the nucleosome

a, Close up view on zinc-finger interactions with the acidic patch. Color code as in Fig 1. b, Protein sequence alignment of human sirtuins. Red boxes and asterisks depict the residues of SIRT6 interacting with the acidic patch of the nucleosome. c, Protein sequence alignment of SIRT6 from different species highlighting the same amino acids as in panel b. Organisms: mm Mus musculus, rn Rattus norvegicus, bt Bos taurus, hs Homo sapiens and cc Castor canadensis. d, Depiction of the 3 arginines of SIRT6 (magenta) interacting with the DNA (grey). e, H2A c-terminal tail (yellow) interacts with SIRT6(magenta). f, SIRT6 binding to H2A or H2AZ containing nucleosomes. Bars show the fraction of residual nucleosomes that did not shift with bound SIRT6 in an electron-mobility shift assay (Supplemental Fig. 10). Bars represent mean ± SD of three biological replicates (shown as dots). ** indicates a statistically significant difference between the fraction of residual H2A and H2A.Z containing nucleosomes (P=0.0016 in paired t-test).

Substituting these four positively charged residues to alanines (SIRT6-4A) significantly weakened the interaction of SIRT6 with the nucleosomes as shown by electrophoretic gel-shift assay (Supplemental Fig. 8a). To test also the activity of SIRT6-4A we first acetylated the H3 tails of the nucleosomes with purified SAGA complex. The mutant protein showed strongly diminished activity in deacetylating H3K9Ac, further corroborating the role of the four basic residues R175, R172, R178 and K170, in binding to the nucleosome (Supplemental Fig. 8b).

All four SIRT6 positive residues are located in a loop that makes a sharp turn, due to a conserved glycine at position 173, connecting two beta strands and penetrates the acidic patch. Interestingly, this loop appears to be flexible in crystal structures of SIRT6 but rendered stable by its association with the acidic patch. Binding to different sections of the acidic patch maintains a strong hold of the zinc finger at the surface of the nucleosome and orients the Rossmann fold domain towards the DNA terminus. This strong hold is maintained throughout the movements of the Rossmann fold domain (Fig. 2).

It is important to note that the four positive residues of SIRT6 that bind the nucleosome acidic patch are conserved among vertebrates but in contrast are not present in other human sirtuins (Fig. 3b,c). This probably reflects the fact that SIRT6 is the only sirtuin predominantly associated with chromatin.

Rossmann fold domain binds nucleosome DNA exit site

Due to its flexibility with respect to the fixed zinc finger domain, the resolution of the Rossmann fold domain is not sufficiently high to distinguish most individual side chains. However secondary structure elements, mostly helices, are clearly visible and guided the precise fitting of the crystal structure of this domain.

We find that three highly conserved arginines from the Rossmann fold domain (R220, R231, R248) bind the phosphate backbone at SHL 6 and 7 of the nucleosomal DNA (Fig. 3d). These interactions are mostly maintained throughout the movement of the DNA terminus with the Rossmann fold domain bound. However, by comparison to dedicated DNA binding proteins such as transcription factors or TBP, the number of observed DNA-protein interactions is low in SIRT6. A biochemical study suggests that the intrinsically disordered C-terminal domain of SIRT6 is necessary for establishing tight interaction of SIRT6 with the nucleosome (Liu et al., 2020) via its ability to bind DNA. Our maps do show an additional isolated protein density that we could not assign with confidence to SIRT6 but associates with the DNA terminus Supplemental Fig. 9a). We propose however that the three arginines are required to correctly position the Rossmann fold, and hence the catalytic site, with respect to the H3 tail. Indeed, applying SAGA-acetylated nucleosomes as substrates, we find that a SIRT6 mutant, where the three arginines are substituted to alanines (SIRT6-3A), has severely diminished H3K9Ac deacetylase activity (Supplemental Fig. 8b). These findings confirm the importance of the three lysines R220, R231 and R248 to the enzymatic function.

Unlike other sirtuins, SIRT6 binds NAD+ in the absence of an acetylated substrate. This is attributed to the fact that crystal structures of SIRT6 describe a rigid helix (α3) where other sirtuins contain a flexible cofactor binding loop (Pan et al., 2011). In our map, we do not observe helix α3 (Supplemental Fig. 9b) and we speculate that in the context of the nucleosome, this helix assumes a flexible loop conformation that like in other sirtuins will bind the cofactor only after stabilization by the acetyl lysine substrate.

In addition, we find that the N-terminal part of the Rossmann fold domain associates with the C-terminal tail of histone H2A at residue K119 (Fig. 3e). This observation suggests that monoubiquitination of H2A at this position, a major histone modification in mammalian cells, could interfere with the binding of SIRT6 or regulate its activity. Furthermore, we used electrophoretic mobility shift assay to follow the binding of SIRT6 to nucleosomes harboring the histone variant H2A.Z which deviates considerably from the canonical H2A in its C-terminal tail but not in other residues that contact SIRT6. Surprisingly, we find that binding of SIRT6 to nucleosomes is mildly (but statistically significant) stronger when H2A.Z is incorporated instead of H2A (Fig. 3f and Supplemental Fig. 10). This finding could be related to SIRT6’s role in repressing gene expression because H2A.Z is enriched in nucleosomes that flank the transcription start site.

SIRT6 is poised to deacetylate the H3 tail

The classical model of the nucleosome posits an elongated and disordered histone H3 N-terminal tail that freely roams and extends away from the core. However recent NMR studies and molecular dynamics predictions show that the tail actually collapses onto the nucleosome core, driven by robust interactions with DNA(Ikebe et al., 2016; Stutzer et al., 2016). The tail adopts multiple structurally heterogeneous, but energetically similar, states. It dynamically transitions between these conformations that all interact with the last superhelix location of nucleosomal DNA (Morrison et al., 2018). We used a published set of 15 predicted H3-tail conformations resulting from end-states of molecular dynamics simulations of the whole nucleosome (Morrison et al., 2018), to analyze whether the position of SIRT6 on the nucleosome serves the enzyme’s role in trapping residues of histone H3 tail. The nucleosome from each molecular dynamics simulation was superposed on our model of SIRT6 bound to a nucleosome with the DNA terminus at the canonical position. To represent the full range of movement of the Rossmann fold domain relative to the DNA we used two endpoints of this movement namely fully bound to DNA and partially dissociated from it (Fig. 4a and b, respectively). We then examined the position of the SIRT6 active site with respect to the predicted paths of the histone H3 tail.

SIRT6 poised to deacetylate lysine residues of H3

a, Side View of SIRT6 bound to the nucleosome with histone H3 K9 residue (red sphere) closest to the SIRT6 active site (blue spheres) from a set of molecular dynamics simulations. b, Side View of SIRT6 bound to the nucleosome with histone H3 K18 residue (orange sphere) closest to the SIRT6 active site from a set of molecular dynamics simulations. c, All H3K9 positions (red spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. d, All H3K18 positions (orange spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. e, Molecular dynamics simulations show that H3 c-terminal tail (blue) can protrude towards SIRT6 in a space formed between the histone octamer and the DNA. H3K27 is shown in red. f, Quantification analysis of H3K27ac bands intensities in deacetylation assay. Bars show percentage of signal detected in Western Blot run with anti-H3K27ac antibodies. Bars represent mean ± SD of three biological replicates (shown as dots). * indicates a statistically significant difference between the 0 min (control) and 60 min (SIRT6 treatment) fraction of acetylated H3K27 (P=0.0396 in one-way paired t-test).

We find that in several H3 conformations, the tail is located in close proximity (<15 Å) to the SIRT6 active site. In some of these cases, H3K9 or H3K18, the main target residues of SIRT6 (Wang et al., 2016), are practically situated in the active site. Furthermore, when analyzing the two endpoints of the Rossmann fold domain shift with respect to the DNA, we observe that in some H3 conformations, this shift draws the active site closer to the target residues (Fig. 4c,d). We conclude that the high-affinity binding of SIRT6 to the nucleosome poises the active site to trapping its main target residues. In addition, the flexibility between the Rossmann fold and Zn-finger domains serves as a fine-tuning mechanism that allows for short-range scanning of the tail in order to locate the acetylated residues. We note also that except H3 no other histone is present at the vicinity of the SIRT6 active site. This gives rise to selectivity of SIRT6 towards the H3 tail.

To shed light on possible benefits offered by SIRT6-mediated DNA unwrapping we conducted molecular dynamic simulations of a nucleosome with bound SIRT6. Strikingly, new H3 tail conformations appear in the simulations that are absent from a control run with nucleosome alone. These pathways have the H3 tail protruding out of the histone core through the space between the DNA and the octamer that is formed via partial unwrapping of the DNA by SIRT6. Importantly, in these additional conformations H3 tail does not need to circle around the DNA to reach SIRT6 and therefore lysines that are closer to the core of H3 (e.g., K27) could also be accessible to the enzyme (Fig. 4e). To lend support to this claim we tested whether SIRT6 can deacetylate residue H3K27 that was first acetylated by SAGA (Figure 4f and Supplemental Fig. 8c). We find that indeed SIRT6 could efficiently deacetylate H3K27Ac, although at a somewhat slower rate than H3K9Ac. We conclude that partial DNA unwrapping by SIRT6 allows H3-tail conformations that make lysines that are close to the core of H3 accessible to the enzyme.

Discussion

Most histone-modifying enzymes, including sirtuin family members such as SIRT1, show significant activity on isolated target peptides and form multiple interactions with the residue to be modified as well as with flanking amino acids to provide selectivity. It is therefore striking that the measured turnover rate for SIRT6 on isolated H3K9 peptides is 1000 slower than for other sirtuins. Moreover, unlike SIRT1, the activity of SIRT6 is dramatically stimulated by binding to its cognate nucleosome substrate.

We solved the structure of SIRT6 bound to a nucleosome and used heterogeneity analysis to uncover the main conformational changes in the complex. Examining these findings in the context of molecular dynamics studies of the nucleosome that produce putative paths of histone tails, shed light on the unique properties of SIRT6.

We find that multiple interactions of SIRT6 with the nucleosome position its active site at the putative path of relatively long-lived conformations of H3 so that residues K9 and K18 of the tail spend significant time very close to, or even at, the enzyme active site. In addition, SIRT6 moves with respect to the nucleosomal DNA allowing it to fine-tune the position of the active site to trap the target residue. Hence it is the binding of SIRT6 to the nucleosome and an intrinsic scanning capability that account for its much higher enzymatic efficiency on nucleosomes.

Importantly, the position of SIRT6 also ensures high selectivity because the space surrounding the active site is devoid of any other tails apart from H3.

The recognition mode of nucleosomal H3 employed by SIRT6 is reminiscent of that used by the PRC1 complex in the specific ubiquitylation of H2A Lys119 (McGinty et al., 2014). Similar to SIRT6, PRC1 requires a nucleosome substrate, showing no activity on an H2A-H2B dimer, and multiple interactions with the nucleosome position an active site in close proximity to its target residue. However, the flexibility and scanning capacity we observed for SIRT6 is lacking in PRC1 perhaps echoing the fact that its target residue is rather stable in the nucleosome compared to the H3 tail.

It is interesting to note that the flexibility of SIRT6 that is at the root of its scanning mechanism was suggested to be, at least in part, responsible for its nearly undetected activity on isolated peptides (Pan et al., 2011). It is only when it is constrained to the proximity of the target residues, through binding to the nucleosome, that this capacity facilitates the function of the enzyme.

A second major conformational change observed in our 3D reconstructions is the detachment of the DNA terminus, with a bound Rossmann fold domain, from the octamer. Our molecular dynamics simulations suggest that the space formed between DNA and octamer allows for additional conformations of the H3 tail that protrudes towards SIRT6. These new pathways do not require circling around the DNA gyre and therefore lysines that are closer to the core of H3 (e.g., K27) can reach in this way the active site of SIRT6. We note that K56 which lies at the core of H3 within a stable helix was also reported to be a target for SIRT6. It is therefore possible that the enzyme employs additional recognition modes of its targets.

Because the position of the SIRT6 active site seems to be well suited for modifying the H3 tail, it stands to reason that other H3 modifying enzymes, such as the GCN5 HAT module of SAGA and ATAC, would place their active site in a position similar to the one of SIRT6 and possibly also displace the DNA from the histone octamer. Gcn5 deviates from SIRT6 in having significant affinity for the H3 tail and therefore might pose a less stringent requirement for precise positioning of the active site or for specific conformations of the H3 tail. Thus, we speculate the Gcn5 recognition mode will be similar to SIRT6 but will also include some sequence specific binding. While this manuscript was under review two studies describing the structure of SIRT6-nucleosome appeared in press (Chio et al., 2023; Wang et al., 2023). The conclusion of these papers regarding the position of SIRT6 on the nucleosome and the unwinding of DNA by the enzyme are similar to our findings. We however dissected in addition the movements of SIRT6 on the nucleosome and analyzed via molecular dynamics the conformations of the H3 tail with respect to the SIRT6 active site. Our results point to the importance of the flexibility between the globular domains of SIRT6 and also explain how SIRT6 can access lysines that are much closer to the histone core than H3K9.

Methods

Human SIRT6 purification

Plasmid (AddGene #28271) was transformed into E. coli strain BL21-RIL. Cells in 1L TB (with 50 µg/mL Kanamycin and 25 µg/mL Chloramphenicol) grew at 37 °C degrees till OD600 reached 0.8 when they were rapidly cooled on ice to 18 °C. Over-expression was induced by the addition 0.5 mM IPTG and cells continued to grow at 18 °C over-night. All subsequent steps were performed at 0–4 °C and solutions were always supplemented with a protease inhibitor cocktail (PMSF, Pepstatin, E-64). Cells were harvested by centrifugation at 8250g for 20 minutes, suspended in buffer A200 (20 mM HepesKOH pH 7.5, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 200 mM NaCl) and treated with lysosyme (1 mg/mL) for 30 minutes on ice. Salt concentration was increased to 0.5 M and cells were then lysed by sonication (6*10sec). Debris were pelleted by centrifugation 45,000g*20minutes. The supernatant was incubated with 0.5 mL of Ni beads (1 mL slurry, Roche), equilibrated in buffer A500, for 1h. The solution was poured into a gravity column (CliniScience). Beads were then washed by 10 mL of buffer A500 with 5 mM imidazole before elution by 2 mL of buffer B300 (20 mM HepesKOH pH 7.2, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 300 mM NaCl) with 150 mM imidazole. The eluted protein was diluted with two volumes of buffer D (20 mM HepesKOH pH 7.0, 5% glycerol v/v, 2 mM beta-mercaptoethanol), loaded on a 1 mL Heparin column and eluted with 0.1 M–1 M salt gradient in buffer D. Fraction were analyzed by SDS-PAGE and those containing SIRT6 were pooled and dialyzed against buffer E (20 mM HepesKOH pH 7.5, 2% glycerol v/v, 0.5 mM TCEP and 150 mM NaCl). Human SIRT6 was then concentrated to 7.8 mg/mL, flash-frozen in liquid N2, and kept in aliquotes at –80 °C.

Cloning of SIRT6 mutants

The region of AddGene plasmid #28271 between SacI and HindIII sites were synthetized by SynBio company with changes in DNA sequence leading to K170A, R172A, R175A, R178A, R220A, R231A and R248A mutations in amino acid sequence. A NheI restriction site was also introduced between the first four and last three mutations without changes in amino acid sequence in order to facilitate cloning. A plasmid expressing SIRT6-7A that incoporates all seven mutations was created via ligation of the synthetic DNA fragment into the original expression plasmid between the SacI and HindIII sites. Fragments of the original expression plasmid corresponding to the regions between HindIII and NheI or NheI and SacI were amplified with PCR and were cloned separately through corresponding restriction sites into the plasmid expressing SIRT6-7A to yield plasmids expressing SIRT6 with the first four mutations (SIRT6-4A) or the last three mutations (SIRT6-3A), respectively.

Octamer reconstitution and nucleosome DNA

Octamers were reconstituted from individual Xenopus leavis (canonical) or human (H2A.Z) histones expressed as inclusion bodies according to the standard protocol (Dyer et al., 2004; Luger et al., 1999). Widom-601 145bp DNA was produced using a plasmid harboring 16 copies of this sequence as described by Dyer et al. (Dyer et al., 2004).

Nucleosome core particle reconstitution

Nucleosomes with 145bp Widom-601 positioning sequence were prepared according to NEB Dilution Assembly Protocol (E5350) (https://international.neb.com/protocols/2012/06/02/dilution-assembly-protocol-e5350) with some modifications as follows: 2.75 µM 145bp Widom-601 DNA was mixed with 2.75 µM canonical or H2A.Z histone octamers in a solution containing 2 M NaCl, 1 mM EDTA, 5 mM beta-mercaptoethanol. The solution was incubated for 30 min at RT and then underwent serial dilutions down to 1.48 M, 1 M, 0.6 M, 0.25 M NaCl with buffer LowSalt (10 mM HepesKOH pH 8.0, 2.5 mM beta-mercaptoethanol). After each dilution the solution was kept at RT for 30 minutes. In order to reduce the final NaCl concentration, nucleosomes were concentrated in 0.5 mL 100 KDa cutoff Amicon up to 100 µL, then diluted 5 times with buffer LowSalt. This step was repeated one more time. Finally, nucleosomes were concentrated to 3–4 µM and analyzed in a 5% native 0.2x TBE polyacrylamide gel to ascertain the quality of the sample and absence of free DNA.

Nucleosome acetylation

Canonical Widom-601 nucleosomes (final concentration 1.8 µM) were mixed with the holo-SAGA complex (final 18 nM) purified from Pichia pastoris cells according to (Papai et al., 2020) in 20 mM HepesKOH pH 7.5, 10% glycerol v/v, 50 mM NaCl, 2 mM MgCl2, 5 mM beta-mercaptoethanol, 0.005% Tween-20, and 0.05 mM Acetyl-CoA. Reaction was incubated on ice for 16 hours. On the next day SAGA was removed via incubation of the mixture with Streptavidin Sepharose beads. In order to reduce the remaining Acetyl-CoA concentration, nucleosomes were concentrated in 0.5 mL 100 KDa cutoff Amicon up to 100 µL, then diluted 5 times with buffer 10 mM HepesKOH pH 8.0, 2.5 mM beta-mercaptoethanol. This step was repeated four more time. Finally, the nucleosomes were concentrated to 3.5 µM.

Binding and gel-shift assays

For nucleosome-binding assays, 100 nM of recombinant nucleosomes (harboring either H2A or H2A.Z histone) were incubated in binding buffer (10 mM HepesKOH pH 7.5, 50 mM NaCl, 0.5 mM TCEP, 6% glycerol v/v) with SIRT6 at 1:2 molar ratio. The reactions were performed on ice for 1h in LoBind tubes (Eppendorf) to minimize protein loss. The reactions were then resolved in a 5% 0.2x TBE gel. The gel was stained with Ethidium Bromide. Bands were quantified with ImageJ.

Deacetylation assay

Non-acetylated or SAGA-acetylated canonical Widom-601 nucleosomes were mixed with SIRT6 (wt, SIRT6-4A or SIRT6-3A) at 100 nM and 40 nM, respectively, in 50 mM HepesKOH pH 7.5, 1 mM DTT, 0.2 mg/mL BSA, 1 mM NAD+. Reactions were set on ice and then incubated at 37 °C for 1 or 2 hours. Reactions were quenched by boiling with 5x Laemmli buffer. The proteins were separated on 15% SDS-PAAG and acetylation was analyzed by Western blot using the following antibodies: for H3K27Ac CellSignaling #8173S diluted 1:1000; for H3K9Ac Abcam Ab4441 1:2500; for H3 CellSignaling #14269S 1:1000.

Grid preparation

SIRT6 was first diluted in buffer (20 mM HepesKOH pH 7.5, 150 mM NaCl, 0.5 mM TCEP, 2% Trehalose). Nucleosome and SIRT6 were then mixed at a 1:2 molar ratio. Final conditions included 1.7 µM Nucleosome and 3.4 µM SIRT6 in buffer (20 mM HepesKOH pH 7.5, 18 mM NaCl, 0.5 mM TCEP). After 1h on ice cross-linking agent glutaraldehyde and the detergent dodecyl-maltoside were added at a final concentration of 0.05% v/v and 0.0125% w/v, respectively. The sample was incubated with the cross-linker for an additional 30–60 minutes. 2 nm carbon foil was floated on 3.5/1 300 mesh Quantifoil grids. 3 µL of sample was deposited onto such grids, blotted 5s with blot force 5 after 60s incubation in a Vitrobot IV (Thermo Fisher) at 6 °C using 100% humidity and flash-frozen in liquid ethane.

cryo-EM data acquisition

Two different datasets were collected. For the first dataset images were acquired on a Cs-corrected Titan Krios G1 (Thermo Fisher) microscope operating at 300 kV in nanoprobe mode using serialEM for automated data collection. Movie frames were recorded on a Gatan K3 direct electron detector after a Quantum LS 967 energy filter using a 20 e-V slit width in zero-loss mode. Images were acquired hardware-binned at a nominal magnification of 81,000, which yielded a pixel size of 0.862 Å. Forty movie frames were recorded at a dose of 1.12 and 1.30 e− per Å2 per frame. The second image dataset was acquired on a Titan Krios G4 microscope operating at 300 kV in nanoprobe mode using EPU for automated data collection. Movie frames were recorded on a Flacon 4i direct electron detector after a Selectris X energy filter using a 10 e-V slit width in zero-loss mode. Images were acquired at nominal magnification of 270,000, which yielded a pixel size of 0.458 Å.

Image analysis

For the first dataset movie frames were aligned, dose-weighted in Warp (Tegunov and Cramer, 2019) together with CTF estimation and particles picking. For the second dataset frame alignment, dose-weighting and CTF estimation was done in cryoSPARC (Punjani et al., 2017), picking was done with crYOLO (Wagner et al., 2019). These datasets were analyzed in cryoSPARC and RELION 3.1 (Zivanov et al., 2018) according to standard protocols. Briefly, images were subjected to reference-free 2D classification in cryoSPARC to remove images with contaminations. Ab-initio reconstruction was performed on the selected images followed by heterogenous and homogenous refinements in cryoSPARC. Structural variability and flexibility were explored with Relion 3.1 using 3D classification without alignment with different T values and cryoDRGN (Zhong et al., 2021) according to standard protocols. Sharpening of the maps was done with DeepEmhancer (Sanchez-Garcia et al., 2021).

Model building and refinement

A composite map was created from the part corresponding to the SIRT6 Rossmann-fold from structure of the first dataset and the SIRT6 Zn-finger plus nucleosome from the structure of the second dataset. Crystal structure of Widom-601 sequence containing nucleosome (PDB: 3LZ0) and human SIRT6 (PDB: 5X16) were rigid-body fitted into this composite cryo-EM structure in Chimera (Pettersen et al., 2004). Flexible-fittings were performed with Isolde (Croll, 2018) in ChimeraX (Goddard et al., 2018) into the cryoDRGN structures. All models were refined in realspace with PHENIX (Liebschner et al., 2019) and Isolde.

Depiction of SIRT6 – nucleosome structures

The following maps are depicted in the figures:

Figure 1. : EMD-16845; Figure 2: b1-cryoDRGN run1 PC1 volume 0, b2-volume 9; e1-cryoDRGN run2 PC1 volume 0, e2-volume 9; Figure 3: a,d-EMD-16845, e-EMD-18497; Figure 4: a-cryoDRGN run2 PC1 volume 0, b-volume 9; Supplemental Figure 4: a,b-EMD-16842, c-EMD-16843, d-EMD-18497; Supplemental Figure 5: a-EMD-16842, c-cryoDRGN run1 PC1 volume 0 and nucleosome void of SIRT6 (Suppl. Fig. 2), d-nucleosome void of SIRT6; Supplemental Figure 9: EMD-18497.

Molecular Dynamics simulations

All molecular dynamics (MD) simulations were performed with the AMBER20 suit of programs, using the ff14SB (Maier et al., 2015) force field with bsc1 DNA corrections (Ivani et al., 2016) and CUFIX ions corrections (Yoo and Aksimentiev, 2018) that improve the description of disordered region conformations. Parameters for the zinc finger were taken from Macchiagodena et al. (Macchiagodena et al., 2020).

The histone tails were added to the cryo-EM structure based on the work described by Armeev et al (Armeev et al., 2021), and protonation states of the protein residues were checked with the propka3 software (Olsson et al., 2011). The system was soaked in a truncated octrahedral TIP3P water box of 20 Å buffer and 0.15 mM of NaCl ions were added to simulate physiological conditions, resulting in a total of 330,000 atoms.

The minimization of the energy was carried out in 5 steps, each one alternated with a 20ps NVT equilibration step at 100K to ensure a very smooth optimization of the starting geometry. Each minimization run included 5,000 steps in steepest descent and 5,000 steps in conjugate gradient, with decreasing restraints on the protein and nucleic atoms from 20kcal/mol to 5kcal/mol that were kept in the associated equilibration step. The last minimization step was released of any constraints and followed by a final NVT equilibration run at 300K. The time step was then switched from 2 to 4fs as allowed by the use of the SHAKE bond length constraints (Kräutler et al., 2001) and the Hydrogen Mass Repartioning algorithm (Hopkins et al., 2015). The latter was applied on the topology using the parmed tool of Amber20. A 40ns equilibration run in NTP was then carried out, followed by a 5µs production. Coordinates were outputted every 1ns during the production run. The temperature was kept constant using the Langevin thermostat with a 2ps-1 collision frequency, and the Berendsen barostat was used to maintain the pressure at 1 atm in the NTP runs. Electrostatic interactions were treated using the Particle Mesh Ewald approach (Darden et al., 1993) with an 8 Å cut-off. Three replicates were performed starting with different velocities at the unconstrained 300K NVT equilibration step, resulting in a total of 15µs of sampling.

A cluster analysis was performed on the MD ensembles based on the DNA and protein RMSD. Conformations were extracted every 5 frames of the trajectories to avoid memory overload and were sorted using the hierarchical agglomerative and average linkage algorithms. 10 clusters were generated by replicate (Supplemental Table 2).

The DNA flexibility analysis was performed on each MD ensemble by using a Principal Component Analysis-based machine learning script quantifying the contribution of each residue to the overall dynamics of the system. Based on the Scikit-learn58 library (Fleetwood et al., 2020), it calculates a covariance matrix from the internal coordinates of the MD ensemble. In this matrix, the eigenmodes showing the highest amplitude are the most implicated in the main fluctuations of the system, and the contribution of each residue is normalized based on these values.

The control system was generated by removing SIRT6 from the initial structure and was submitted to the same protocol.

Data availability

The cryo-EM maps have been deposited in the Electron Microscopy Data Bank (EMDB) under accession codes EMD-16842 (High-resolution on the nucleosome), EMD-16843 (SIRT6-nucleosome refined for SIRT6), EMD-18497 (Tracing the H2A tail) and EMD-16845 (composite map). CryoDRGN maps are available upon request. The model coordinates for SIRT6-nucleosome were deposited in the Protein Data Bank (PDB) under the accession code 8OF4.

All MD input files are publicly available on Github: https://github.com/emmanuellebignon/NCP-SIRT6

Supporting information

eLife Assessment

This manuscript provides a useful reconstruction of the structure of the sirtuin-class histone deacetylase Sirt6 bound to a nucleosome based on cryo-EM observations, and additional characterization of the flexibility of the histone tails in the complex based on molecular dynamics simulations. While similar structures have recently been published, this solid study supports the conclusions of those papers and also includes new insights into the potential dynamics of Sirt6 bound to a nucleosome, insights that help explain its substrate specificity. Unfortunately, the authors do not mention the other recent publications until the end of their Discussion, and therefore provide little opportunity for comparison or context for the results presented.

Reviewer #1 (Public Review):

Smirnova et al. present a cryo-EM structure of a nucleosome-SIRT6 complex to understand how the histone deacetylase SIRT6 deacetylates the N-terminal tail of histone H3. The authors obtained the structure at sub-4 Å resolution and can visualize how interactions between the nucleosome and SIRT6 position SIRT6 to allow for H3 tail deacetylation. Through additional conformational analysis of their cryo-EM data, they reveal that SIRT6 positioning is flexible on the nucleosome surface, and this could accommodate the targeting of certain H3 tail residues. This work is significant as it represents the visualization of a histone deacetylase on its native nucleosomal target and reveals how substrate specificity is achieved. Importantly, it should be noted that recently two additional structures of the nucleosome-SIRT6 complex were already published. Therefore, Smirnova et al. confirm and complement these previous findings. Additionally, Smirnova et al. expand our understanding of the structural flexibility of SIRT6 on the nucleosome and clarify that SIRT6 also shows histone deacetylase activity on H3K27Ac.

Reviewer #2 (Public Review):

Smirnova et al. present a cryo-EM structure of human SIRT6 bound to a nucleosome as well as the results from molecular dynamics simulations. The results show that the combined conformational flexibilities of SIRT6 and the N-terminal tail of histone H3 limit the residues with access to the active site, partially explaining the substrate specificity of this sirtuin-class histone deacetylase. Two other groups have recently published cryo-EM structures of SIRT6:nucleosome complexes; this manuscript confirms and complements these previous findings, with the addition of some novel insights into the role of structural flexibility in substrate selection.

Author Response

The following is the authors’ response to the previous reviews.

We thank the reviewers for their remarks. Please find our detailed answers bellow.

1. The authors' continued refusal to acknowledge the other reports before the final sentence of the Discussion, which has been pointed out in two previous rounds of review as a major flaw, detracts from the manuscript significantly.

We now acknowledge and discuss the other SIRT6-nucleosome reports in the introduction as requested by the reviewer.

1. While some of the grammatical errors in previous versions have been corrected, many remain, especially in the Methods section

We corrected the remaining grammatical errors.

1. Multiple statements of fact not supported by data shown in this work continue to lack appropriate references.

We added references where facts were not supported by our data.

Version 4

Abstract

Abstract

Sirtuin 6 (SIRT6) is an NAD+ dependent histone H3 deacetylase that is prominently found associated with chromatin, attenuates transcriptionally active promoters and regulates DNA repair, metabolic homeostasis and lifespan. Unlike other sirtuins, it has low affinity to free histone tails but demonstrates strong binding to nucleosomes. It is poorly understood how SIRT6 docking on nucleosomes stimulates its histone deacetylation activity. Here we present the structure of human SIRT6 bound to a nucleosome determined by cryogenic electron microscopy. The zinc finger domain of SIRT6 associates tightly with the acidic patch of the nucleosome through multiple arginine anchors. The Rossmann fold domain binds to the terminus of the looser DNA half of the nucleosome, detaching two turns of the DNA from the histone octamer and placing the NAD+ binding pocket close to the DNA exit site. This domain shows flexibility with respect to the fixed zinc finger and moves with, but also relative to, the unwrapped DNA terminus. We apply molecular dynamics simulations of the histone tails in the nucleosome to show that in this mode of interaction, the active site of SIRT6 is perfectly poised to catalyze deacetylation of the H3 histone tail and that the partial unwrapping of the DNA allows even lysines close to the H3 core to reach the enzyme.

Introduction

The state of lysine acetylation in histones, as well as in several other proteins, is determined by the balancing act of acetyltransferases and deacetylases. Sirtuin is an evolutionarily conserved family of lysine deacetylases that comprises seven members in humans and plays an important role in regulating numerous physiological and pathological processes including aging, cancer, inflammation and metabolism (Cen et al., 2011; Vitiello et al., 2017; Watroba et al., 2017). Unlike other lysine deacetylases sirtuins carry out their enzymatic reaction by transferring the acetyl group to an essential NAD+ cofactor before acetylating the target protein (Imai et al., 2000).

Human Sirtuin 6 (SIRT6) is a multitasking enzyme implicated in diverse pathways. Through its histone H3 deacetylation at gene promoters, it is a transcription repressor attenuating inter-alia ribosome biogenesis (Sebastian et al., 2012), glycolysis (Taylor et al., 2022) and the expression of various transcription factors, including nuclear factor κB (NF-κB) (Kawahara et al., 2009) and hypoxia-inducible factor-1α (HIF-1α) (Zhong et al., 2010). Its H3K9 deacetylation of telomeric chromatin maintains telomere integrity (Michishita et al., 2008). In addition, via its second enzymatic activity, ADP-ribosylation of PARP1 (Mao et al., 2011) and by stabilizing DNA-PK on chromatin (McCord et al., 2009), SIRT6 also fosters DNA double-strand break repair.

These activities position SIRT6 as a key regulator of longevity as indeed shown by several lines of evidence. In mice, knockout of SIRT6 leads to a shortened lifespan and a premature aging-like phenotype (Kanfi et al., 2012). Similar observations were made in monkeys deficient for the enzyme (Zhang et al., 2018). Conversely, overexpression of SIRT6 extends lifespan and diminishes frailty in mice (Roichman et al., 2021). Furthermore, a study with several rodent species found a positive correlation between SIRT6 ability to promote DNA double-strand break repair and longevity - in other words, more active SIRT6 leads to longer lifespan (Tian et al., 2019).

The development of cancer can also be regulated by SIRT6. However, whereas in some malignancies SIRT6 plays a role as a tumor suppressor (Marquardt et al., 2013; Min et al., 2012; Sebastian et al., 2012), in others it acts as an oncogene (Bauer et al., 2012; Khongkow et al., 2013; Liu et al., 2013). The molecular mechanisms underlying SIRT6 effects on longevity or on the two contradictory roles in cancer development are not well understood, limiting the search for medically useful drugs that target the enzyme.

SIRT6 is a 37 KDa protein composed of two globular parts, a Rossmann fold domain that harbors the NAD+ binding site and a zinc finger domain (Pan et al., 2011). A striking difference between SIRT6 and other members of the sirtuin family is the presence of a splayed zinc finger and the lack of a stabilizing helix bundle linking the two globular domains. This absence probably endows SIRT6 with a unique flexibility between the domains (Supplemental Fig. 1).

These structural differences are manifested in function. Unlike other sirtuins, SIRT6 has very low activity on acetylated free H3 histones or histone peptides (Jiang et al., 2013). This seemed to contradict findings concerning the function of SIRT6, in particular data demonstrating its activity in vivo as H3 deacetylase (Michishita et al., 2008) as well as the fact that SIRT6 is nuclear and found mainly associated with chromatin (Michishita et al., 2005). The conundrum was settled when it was revealed that nucleosomes, the in vivo cognate substrate of SIRT6, dramatically enhance the enzyme activity towards acetylated H3 histones (Gil et al., 2013). Further biochemical analysis showed that SIRT6 binds nucleosomes with high affinity via multiple interacting sites (Liu et al., 2020).

However, a mechanistic understanding of SIRT6 binding to nucleosomes and how this association promotes higher enzyme efficacy is currently lacking.

Here we present the structure of human SIRT6 bound to the nucleosome. We identify interactions between the zinc finger domain and the acidic patch of the nucleosome as well as between the Rossmann fold and the DNA terminus, positioning the NAD+ binding pocket close to the nucleosome DNA exit site and partially detaching the DNA from the histone octamer. While this manuscript was under review two studies describing the structure of SIRT6-nucleosome appeared in press (Chio et al., 2023; Wang et al., 2023). The conclusion of these papers regarding the position of SIRT6 on the nucleosome and the unwinding of DNA by the enzyme are similar to our findings. We however use in addition molecular dynamics simulations of the nucleosome to show that in this position the enzyme active site is perfectly poised to bind the H3 tail for deacetylation and that due to the unwrapping of the DNA even lysines close to the core of H3 can reach the enzyme. Furthermore, we find conformational changes, facilitated by the flexibility in SIRT6 between the zinc finger and the Rossmann fold, that could serve to fine-tune the position of the active site with respect to the H3 tail in order to trap its target acetylated residues.

Results

Overall structure

To investigate how nucleosomes stimulate the activity of SIRT6 we prepared nucleosome with the Widom-601 positioning sequence (Lowary and Widom, 1998) and reconstituted its complex with full-length recombinant human SIRT6 that was overexpressed in and purified from E. coli. The complex was stabilized via mild crosslinking before its deposition on cryo-EM grids.

We obtained a single particle cryogenic electron microscopy structure showing the nucleosome at 2.9 Å overall resolution and a protein density bound to its disc-like surface (Fig. 1a and Supplemental Fig. 2 and 3). SIRT6 could be fitted into this density with its zinc finger domain (resolved at 3.0 Å resolution), associated with the acidic patch of the nucleosome and its Rossmann fold domain (4.0–6.0 Å resolution) interacting with the DNA terminus at superhelix locations (SHL) 6 and 7 (Fig. 1b and Supplemental Fig. 4). The active site of SIRT6 is thus positioned just above where the H3 tail enters the nucleosome disc between the two DNA gyres. The high-resolution at the nucleosome allows discerning purines from pyrimidines. We could therefore distinguish between the two pseudosymmetric orientations of the Widom-601 DNA (Supplemental Fig. 5a). We find that SIRT6 is bound predominantly at the “right” TA-poor arm of the Widom-601 sequence where DNA interactions with the histone octamer are looser (Ngo et al., 2015). We detect however a small population with a second SIRT6 bound at the “left” TA-rich tight arm (Supplemental Fig. 2).

Structure of human SIRT6 in complex with the nucleosome

a, Front (top) and side (bottom) views of a composite cryo-EM reconstruction of human SIRT6-nucleosome. Maps from focused refinements of SIRT6 (magenta – Rossmann fold and Zn-finger domains labelled) and the nucleosome (H2A – yellow; H2B – orange; H3 – blue; H4 – green; DNA – light and dim grey). b, Corresponding views of the atomic model of the complex.

Conformational heterogeneity analyses in cryoDRGN (Zhong et al., 2021) reveal two principle large movements within the structure (Fig. 2a,d and Supplemental Fig. 6).

Flexibility of the SIRT6 Rossmann fold

a, UMAP projection of the latent embeddings of a subset of SIRT6-nucleosome particles representing the concerted movement of the Rossmann fold and the DNA terminus. b, Structural representation of the two endpoints of the latent embeddings shown in panel a. c, Overlay of the DNA ends and the SIRT6 Rossmann fold of two endpoints shown in panel b. Arrows show their displacements between the two endpoints. d, UMAP projection of the latent embeddings of another subset of SIRT6-nucleosome particles representing the movement of the Rossmann fold with respect to the nucleosomal DNA. e, Structural representation of the two endpoints of the latent embeddings shown in panel b. f, Overlay of the DNA ends and the SIRT6 Rossmann fold of two endpoints shown in panel e. Arrow show the displacement of the Rossmann fold between the two endpoints.

First, we find that where the SIRT6 Rossmann fold binds, DNA is pulled away and detached from the histone octamer (Fig. 2b). Particles greatly vary in the extent of this shift that therefore probably represents a continuous movement of the DNA (compare Fig. 2b1 with 2b2). The Rossmann fold moves with and is attached to the DNA whereas the zinc finger domain remains fixed at the acidic patch of the nucleosome (Fig. 2c). In the full extent of the DNA movement, two turns from SHL 6 and 7 are detached from the histone octamer. This alters the path of the terminus DNA by 37° and breaks DNA interactions with histones H2A, H2B, and H3 (Supplemental Fig. 5b). Such a large change in DNA trajectory is reminiscent of the chromatin remodelling ATPases or pioneer transcription factors binding to nucleosome but was not observed in other histone modifiers (Dodonova et al., 2020; Farnung et al., 2017). Importantly, the DNA terminus without bound SIRT6 or free nucleosomes in our dataset did not show any deviation from the canonical conformations (Supplemental Fig. 5c). In addition, our MD simulations indicate higher flexibility of the DNA termini when SIRT6 is bound to the nucleosome (Supplemental Fig. 7). All together these findings strongly suggest that the DNA movement we observed is coupled to SIRT6 interactions.

Second, in particles where the SIRT6-bound DNA remains in its canonical position, we observe shifts by up to 20 Å in the position of the Rossmann fold domain of SIRT6, partially loosening its association with the DNA (Fig. 2e,f). We show that this shift could be involved in fine tuning the position of the active site relative to the H3 tail.

These conformational changes are made possible by a flexible hinge region between the two domains of SIRT6, which is unique to this member of the sirtuins family.

Zinc finger domain binds the acidic patch

An important feature of the nucleosome surface is a cluster of acidic residues (E56, E61, E64, D90, E91, E92 of H2A and E102, E110 of H2B) that forms a negatively charged ‘acidic patch’ and takes the shape of a narrow groove. It serves as a docking site for many nucleosome-binding proteins. Four side chains of the SIRT6 zinc-finger domain extend into different sections of the acidic patch groove and interact with the negatively charged residues. R175 packs against H2B L103 and serves as a classical “arginine anchor motif” (McGinty and Tan, 2016), binding to the acidic triad of E61, D90 and E92 from H2A. SIRT6 presents two other arginines, R172, to contact E56 and Q24 of H2A as well as Q44 and E110 of H2B and R178 to contact H2A E91 and E102 from H2B. Finally, K170 interacts with E64 and N68 from H2A (Fig. 3a).

Binding of SIRT6 to the nucleosome

a, Close up view on zinc-finger interactions with the acidic patch. Color code as in Fig 1. b, Protein sequence alignment of human sirtuins. Red boxes and asterisks depict the residues of SIRT6 interacting with the acidic patch of the nucleosome. c, Protein sequence alignment of SIRT6 from different species highlighting the same amino acids as in panel b. Organisms: mm Mus musculus, rn Rattus norvegicus, bt Bos taurus, hs Homo sapiens and cc Castor canadensis. d, Depiction of the 3 arginines of SIRT6 (magenta) interacting with the DNA (grey). e, H2A c-terminal tail (yellow) interacts with SIRT6(magenta). f, SIRT6 binding to H2A or H2AZ containing nucleosomes. Bars show the fraction of residual nucleosomes that did not shift with bound SIRT6 in an electron-mobility shift assay (Supplemental Fig. 10). Bars represent mean ± SD of three biological replicates (shown as dots). ** indicates a statistically significant difference between the fraction of residual H2A and H2A.Z containing nucleosomes (P=0.0016 in paired t-test).

Substituting these four positively charged residues to alanines (SIRT6-4A) significantly weakened the interaction of SIRT6 with the nucleosomes as shown by electrophoretic gel-shift assay (Supplemental Fig. 8a). To test also the activity of SIRT6-4A we first acetylated the H3 tails of the nucleosomes with purified SAGA complex. The mutant protein showed strongly diminished activity in deacetylating H3K9Ac, further corroborating the role of the four basic residues R175, R172, R178 and K170, in binding to the nucleosome (Supplemental Fig. 8b).

All four SIRT6 positive residues are located in a loop that makes a sharp turn, due to a conserved glycine at position 173, connecting two beta strands and penetrates the acidic patch. Interestingly, this loop appears to be flexible in crystal structures of SIRT6 but rendered stable by its association with the acidic patch. Binding to different sections of the acidic patch maintains a strong hold of the zinc finger at the surface of the nucleosome and orients the Rossmann fold domain towards the DNA terminus. This strong hold is maintained throughout the movements of the Rossmann fold domain (Fig. 2).

It is important to note that the four positive residues of SIRT6 that bind the nucleosome acidic patch are conserved among vertebrates but in contrast are not present in other human sirtuins (Fig. 3b,c). This probably reflects the fact that SIRT6 is the only sirtuin predominantly associated with chromatin.

Rossmann fold domain binds nucleosome DNA exit site

Due to its flexibility with respect to the fixed zinc finger domain, the resolution of the Rossmann fold domain is not sufficiently high to distinguish most individual side chains. However secondary structure elements, mostly helices, are clearly visible and guided the precise fitting of the crystal structure of this domain.

We find that three highly conserved arginines from the Rossmann fold domain (R220, R231, R248) bind the phosphate backbone at SHL 6 and 7 of the nucleosomal DNA (Fig. 3d). These interactions are mostly maintained throughout the movement of the DNA terminus with the Rossmann fold domain bound. However, by comparison to dedicated DNA binding proteins such as transcription factors or TBP, the number of observed DNA-protein interactions is low in SIRT6. A biochemical study suggests that the intrinsically disordered C-terminal domain of SIRT6 is necessary for establishing tight interaction of SIRT6 with the nucleosome (Liu et al., 2020) via its ability to bind DNA. Our maps do show an additional isolated protein density that we could not assign with confidence to SIRT6 but associates with the DNA terminus Supplemental Fig. 9a). We propose however that the three arginines are required to correctly position the Rossmann fold, and hence the catalytic site, with respect to the H3 tail. Indeed, applying SAGA-acetylated nucleosomes as substrates, we find that a SIRT6 mutant, where the three arginines are substituted to alanines (SIRT6-3A), has severely diminished H3K9Ac deacetylase activity (Supplemental Fig. 8b). These findings confirm the importance of the three lysines R220, R231 and R248 to the enzymatic function.

Unlike other sirtuins, SIRT6 binds NAD+ in the absence of an acetylated substrate (Imai et al., 2000). This is attributed to the fact that crystal structures of SIRT6 describe a rigid helix (α3) where other sirtuins contain a flexible cofactor binding loop (Pan et al., 2011). In our map, we do not observe helix α3 (Supplemental Fig. 9b) and we speculate that in the context of the nucleosome, this helix assumes a flexible loop conformation that like in other sirtuins will bind the cofactor only after stabilization by the acetyl lysine substrate.

In addition, we find that the N-terminal part of the Rossmann fold domain associates with the C-terminal tail of histone H2A at residue K119 (Fig. 3e). This observation suggests that monoubiquitination of H2A at this position, a major histone modification in mammalian cells, could interfere with the binding of SIRT6 or regulate its activity. Furthermore, we used electrophoretic mobility shift assay to follow the binding of SIRT6 to nucleosomes harboring the histone variant H2A.Z which deviates considerably from the canonical H2A in its C-terminal tail but not in other residues that contact SIRT6. Surprisingly, we find that binding of SIRT6 to nucleosomes is mildly (but statistically significant) stronger when H2A.Z is incorporated instead of H2A (Fig. 3f and Supplemental Fig. 10). This finding could be related to SIRT6’s role in repressing gene expression because H2A.Z is enriched in nucleosomes that flank the transcription start site.

SIRT6 is poised to deacetylate the H3 tail

The classical model of the nucleosome posits an elongated and disordered histone H3 N-terminal tail that freely roams and extends away from the core. However recent NMR studies and molecular dynamics predictions show that the tail actually collapses onto the nucleosome core, driven by robust interactions with DNA(Ikebe et al., 2016; Stutzer et al., 2016). The tail adopts multiple structurally heterogeneous, but energetically similar, states. It dynamically transitions between these conformations that all interact with the last superhelix location of nucleosomal DNA (Morrison et al., 2018). We used a published set of 15 predicted H3-tail conformations resulting from end-states of molecular dynamics simulations of the whole nucleosome (Morrison et al., 2018), to analyze whether the position of SIRT6 on the nucleosome serves the enzyme’s role in trapping residues of histone H3 tail. The nucleosome from each molecular dynamics simulation was superposed on our model of SIRT6 bound to a nucleosome with the DNA terminus at the canonical position. To represent the full range of movement of the Rossmann fold domain relative to the DNA we used two endpoints of this movement namely fully bound to DNA and partially dissociated from it (Fig. 4a and b, respectively). We then examined the position of the SIRT6 active site with respect to the predicted paths of the histone H3 tail.

SIRT6 poised to deacetylate lysine residues of H3

a, Side View of SIRT6 bound to the nucleosome with histone H3 K9 residue (red sphere) closest to the SIRT6 active site (blue spheres) from a set of molecular dynamics simulations. b, Side View of SIRT6 bound to the nucleosome with histone H3 K18 residue (orange sphere) closest to the SIRT6 active site from a set of molecular dynamics simulations. c, All H3K9 positions (red spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. d, All H3K18 positions (orange spheres) in close proximity (<15 Å) to the SIRT6 active site (blue) taken from a set of 15 molecular dynamics simulations and depicted on the surface view of SIRT6 bound to nucleosome. e, Molecular dynamics simulations show that H3 c-terminal tail (blue) can protrude towards SIRT6 in a space formed between the histone octamer and the DNA. H3K27 is shown in red. f, Quantification analysis of H3K27ac bands intensities in deacetylation assay. Bars show percentage of signal detected in Western Blot run with anti-H3K27ac antibodies. Bars represent mean ± SD of three biological replicates (shown as dots). * indicates a statistically significant difference between the 0 min (control) and 60 min (SIRT6 treatment) fraction of acetylated H3K27 (P=0.0396 in one-way paired t-test).

We find that in several H3 conformations, the tail is located in close proximity (<15 Å) to the SIRT6 active site. In some of these cases, H3K9 or H3K18, the main target residues of SIRT6 (Wang et al., 2016), are practically situated in the active site. Furthermore, when analyzing the two endpoints of the Rossmann fold domain shift with respect to the DNA, we observe that in some H3 conformations, this shift draws the active site closer to the target residues (Fig. 4c,d). We conclude that the high-affinity binding of SIRT6 to the nucleosome poises the active site to trapping its main target residues. In addition, the flexibility between the Rossmann fold and Zn-finger domains serves as a fine-tuning mechanism that allows for short-range scanning of the tail in order to locate the acetylated residues. We note also that except H3 no other histone is present at the vicinity of the SIRT6 active site. This gives rise to selectivity of SIRT6 towards the H3 tail.

To shed light on possible benefits offered by SIRT6-mediated DNA unwrapping we conducted molecular dynamic simulations of a nucleosome with bound SIRT6. Strikingly, new H3 tail conformations appear in the simulations that are absent from a control run with nucleosome alone. These pathways have the H3 tail protruding out of the histone core through the space between the DNA and the octamer that is formed via partial unwrapping of the DNA by SIRT6. Importantly, in these additional conformations H3 tail does not need to circle around the DNA to reach SIRT6 and therefore lysines that are closer to the core of H3 (e.g., K27) could also be accessible to the enzyme (Fig. 4e). To lend support to this claim we tested whether SIRT6 can deacetylate residue H3K27 that was first acetylated by SAGA (Figure 4f and Supplemental Fig. 8c). We find that indeed SIRT6 could efficiently deacetylate H3K27Ac, although at a somewhat slower rate than H3K9Ac. We conclude that partial DNA unwrapping by SIRT6 allows H3-tail conformations that make lysines that are close to the core of H3 accessible to the enzyme.

Discussion

Most histone-modifying enzymes, including sirtuin family members such as SIRT1, show significant activity on isolated target peptides and form multiple interactions with the residue to be modified as well as with flanking amino acids to provide selectivity (Zhao et al., 2013). It is therefore striking that the measured turnover rate for SIRT6 on isolated H3K9 peptides is 1000 slower than for other sirtuins (Jiang et al., 2013). Moreover, unlike SIRT1, the activity of SIRT6 is dramatically stimulated by binding to its cognate nucleosome substrate (Gil et al., 2013).

We solved the structure of SIRT6 bound to a nucleosome and used heterogeneity analysis to uncover the main conformational changes in the complex. Examining these findings in the context of molecular dynamics studies of the nucleosome that produce putative paths of histone tails, shed light on the unique properties of SIRT6.

We find that multiple interactions of SIRT6 with the nucleosome position its active site at the putative path of relatively long-lived conformations of H3 so that residues K9 and K18 of the tail spend significant time very close to, or even at, the enzyme active site. In addition, SIRT6 moves with respect to the nucleosomal DNA allowing it to fine-tune the position of the active site to trap the target residue. Hence it is the binding of SIRT6 to the nucleosome and an intrinsic scanning capability that account for its much higher enzymatic efficiency on nucleosomes. Importantly, the position of SIRT6 also ensures high selectivity because the space surrounding the active site is devoid of any other tails apart from H3.

The recognition mode of nucleosomal H3 employed by SIRT6 is reminiscent of that used by the PRC1 complex in the specific ubiquitylation of H2A Lys119 (McGinty et al., 2014). Like SIRT6, PRC1 requires a nucleosome substrate, showing no activity on an H2A-H2B dimer, and multiple interactions with the nucleosome position an active site in close proximity to its target residue. However, the flexibility and scanning capacity we observed for SIRT6 is lacking in PRC1 perhaps echoing the fact that its target residue is rather stable in the nucleosome compared to the H3 tail.

It is interesting to note that the flexibility of SIRT6 that is at the root of its scanning mechanism was suggested to be, at least in part, responsible for its nearly undetected activity on isolated peptides (Pan et al., 2011). It is only when it is constrained to the proximity of the target residues, through binding to the nucleosome, that this capacity facilitates the function of the enzyme.

A second major conformational change observed in our 3D reconstructions is the detachment of the DNA terminus, with a bound Rossmann fold domain, from the octamer. Our molecular dynamics simulations suggest that the space formed between DNA and octamer allows for additional conformations of the H3 tail that protrudes towards SIRT6. These new pathways do not require circling around the DNA gyre and therefore lysines that are closer to the core of H3 (e.g., K27) can reach in this way the active site of SIRT6. We note that K56 which lies at the core of H3 within a stable helix was also reported to be a target for SIRT6. It is therefore possible that the enzyme employs additional recognition modes of its targets.

Because the position of the SIRT6 active site seems to be well suited for modifying the H3 tail, it stands to reason that other H3 modifying enzymes, such as the GCN5 HAT module of SAGA and ATAC, would place their active site in a position similar to the one of SIRT6 and possibly also displace the DNA from the histone octamer. Gcn5 deviates from SIRT6 in having significant affinity for the H3 tail and therefore might pose a less stringent requirement for precise positioning of the active site or for specific conformations of the H3 tail. Thus, we speculate the Gcn5 recognition mode will be similar to SIRT6 but will also include some sequence specific binding.

Methods

Human SIRT6 purification

Plasmid (AddGene #28271) was transformed into E. coli strain BL21-RIL. Cells in 1L TB (with 50 µg/mL Kanamycin and 25 µg/mL Chloramphenicol) grew at 37 °C degrees till OD600 reached 0.8 when they were rapidly cooled on ice to 18 °C. Over-expression was induced by the addition 0.5 mM IPTG and cells continued to grow at 18 °C over-night. All subsequent steps were performed at 0–4 °C and solutions were always supplemented with a protease inhibitor cocktail (PMSF, Pepstatin, E-64). Cells were harvested by centrifugation at 8250g for 20 minutes, suspended in buffer A200 (20 mM HepesKOH pH 7.5, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 200 mM NaCl) and treated with lysosyme (1 mg/mL) for 30 minutes on ice. Salt concentration was increased to 0.5 M and cells were then lysed by sonication (6*10sec). Debris was pelleted by centrifugation 45,000g*20minutes. The supernatant was incubated with 0.5 mL of Ni beads (1 mL slurry, Roche), equilibrated in buffer A500, for 1h. The solution was poured into a gravity column (CliniScience). Beads were then washed with 10 mL of buffer A500 with 5 mM imidazole before elution by 2 mL of buffer B300 (20 mM HepesKOH pH 7.2, 5% glycerol v/v, 2 mM beta-mercaptoethanol and 300 mM NaCl) with 150 mM imidazole. The eluted protein was diluted with two volumes of buffer D (20 mM HepesKOH pH 7.0, 5% glycerol v/v, 2 mM beta-mercaptoethanol), loaded on a 1 mL Heparin column and eluted with 0.1 M–1 M salt gradient in buffer D. Fraction were analyzed by SDS-PAGE and those containing SIRT6 were pooled and dialyzed against buffer E (20 mM HepesKOH pH 7.5, 2% glycerol v/v, 0.5 mM TCEP and 150 mM NaCl). Human SIRT6 was then concentrated to 7.8 mg/mL, flash-frozen in liquid N2, and kept in aliquotes at –80 °C.

Cloning of SIRT6 mutants

The region of AddGene plasmid #28271 between SacI and HindIII sites was synthetized by SynBio company with changes in DNA sequence leading to K170A, R172A, R175A, R178A, R220A, R231A and R248A mutations in amino acid sequence. A NheI restriction site was also introduced between the first four and last three mutations without changes in amino acid sequence in order to facilitate cloning. A plasmid expressing SIRT6-7A that incorporates all seven mutations was created via ligation of the synthetic DNA fragment into the original expression plasmid between the SacI and HindIII sites. Fragments of the original expression plasmid corresponding to the regions between HindIII and NheI or NheI and SacI were amplified with PCR and were cloned separately through corresponding restriction sites into the plasmid expressing SIRT6-7A to yield plasmids expressing SIRT6 with the first four mutations (SIRT6-4A) or the last three mutations (SIRT6-3A), respectively.

Octamer reconstitution and nucleosome DNA

Octamers were reconstituted from individual Xenopus leavis (canonical) or human (H2A.Z) histones expressed as inclusion bodies according to the standard protocol (Dyer et al., 2004; Luger et al., 1999). Widom-601 145bp DNA was produced using a plasmid harboring 16 copies of this sequence as described by Dyer et al. (Dyer et al., 2004).

Nucleosome core particle reconstitution

Nucleosomes with 145bp Widom-601 positioning sequence were prepared according to NEB Dilution Assembly Protocol (E5350) (https://international.neb.com/protocols/2012/06/02/dilution-assembly-protocol-e5350) with some modifications as follows: 2.75 µM 145bp Widom-601 DNA was mixed with 2.75 µM canonical or H2A.Z histone octamers in a solution containing 2 M NaCl, 1 mM EDTA, 5 mM beta-mercaptoethanol. The solution was incubated for 30 min at RT and then underwent serial dilutions down to 1.48 M, 1 M, 0.6 M, 0.25 M NaCl with buffer LowSalt (10 mM HepesKOH pH 8.0, 2.5 mM beta-mercaptoethanol). After each dilution the solution was kept at RT for 30 minutes. In order to reduce the final NaCl concentration, nucleosomes were concentrated in 0.5 mL 100 KDa cutoff Amicon up to 100 µL, then diluted 5 times with buffer LowSalt. This step was repeated one more time. Finally, nucleosomes were concentrated to 3–4 µM and analyzed in a 5% native 0.2x TBE polyacrylamide gel to ascertain the quality of the sample and absence of free DNA.

Nucleosome acetylation

Canonical Widom-601 nucleosomes (final concentration 1.8 µM) were mixed with the holo-SAGA complex (final 18 nM) purified from Pichia pastoris cells according to (Papai et al., 2020) in 20 mM HepesKOH pH 7.5, 10% glycerol v/v, 50 mM NaCl, 2 mM MgCl2, 5 mM beta-mercaptoethanol, 0.005% Tween-20, and 0.05 mM Acetyl-CoA. Reaction was incubated on ice for 16 hours. On the next day SAGA was removed via incubation of the mixture with Streptavidin Sepharose beads. In order to reduce the remaining Acetyl-CoA concentration, nucleosomes were concentrated in 0.5 mL 100 KDa cutoff Amicon up to 100 µL, then diluted 5 times with buffer 10 mM HepesKOH pH 8.0, 2.5 mM beta-mercaptoethanol. This step was repeated four more times. Finally, the nucleosomes were concentrated to 3.5 µM.

Binding and gel-shift assays

For nucleosome-binding assays, 100 nM of recombinant nucleosomes (harboring either H2A or H2A.Z histone) were incubated in binding buffer (10 mM HepesKOH pH 7.5, 50 mM NaCl, 0.5 mM TCEP, 6% glycerol v/v) with SIRT6 at 1:2 molar ratio. The reactions were performed on ice for 1h in LoBind tubes (Eppendorf) to minimize protein loss. The reactions were then resolved in a 5% 0.2x TBE gel. The gel was stained with Ethidium Bromide. Bands were quantified with ImageJ.

Deacetylation assay

Non-acetylated or SAGA-acetylated canonical Widom-601 nucleosomes were mixed with SIRT6 (wt, SIRT6-4A or SIRT6-3A) at 100 nM and 40 nM, respectively, in 50 mM HepesKOH pH 7.5, 1 mM DTT, 0.2 mg/mL BSA, 1 mM NAD+. Reactions were set on ice and then incubated at 37 °C for 1 or 2 hours. Reactions were quenched by boiling with 5x Laemmli buffer. The proteins were separated on 15% SDS-PAAG and acetylation was analyzed by Western blot using the following antibodies: for H3K27Ac CellSignaling #8173S diluted 1:1000; for H3K9Ac Abcam Ab4441 1:2500; for H3 CellSignaling #14269S 1:1000.

Grid preparation

SIRT6 was first diluted in buffer (20 mM HepesKOH pH 7.5, 150 mM NaCl, 0.5 mM TCEP, 2% Trehalose). Nucleosome and SIRT6 were then mixed at a 1:2 molar ratio. Final conditions included 1.7 µM Nucleosome and 3.4 µM SIRT6 in buffer (20 mM HepesKOH pH 7.5, 18 mM NaCl, 0.5 mM TCEP). After 1h on ice cross-linking agent glutaraldehyde and the detergent dodecyl-maltoside were added at a final concentration of 0.05% v/v and 0.0125% w/v, respectively. The sample was incubated with the cross-linker for an additional 30–60 minutes. 2 nm carbon foil was floated on 3.5/1 300 mesh Quantifoil grids. 3 µL of sample was deposited onto such grids, blotted 5s with blot force 5 after 60s incubation in a Vitrobot IV (Thermo Fisher) at 6 °C using 100% humidity and flash-frozen in liquid ethane.

cryo-EM data acquisition

Two different datasets were collected. For the first dataset images were acquired on a Cs-corrected Titan Krios G1 (Thermo Fisher) microscope operating at 300 kV in nanoprobe mode using serialEM for automated data collection. Movie frames were recorded on a Gatan K3 direct electron detector after a Quantum LS 967 energy filter using a 20 e-V slit width in zero-loss mode. Images were acquired hardware-binned at a nominal magnification of 81,000, which yielded a pixel size of 0.862 Å. Forty movie frames were recorded at a dose of 1.12 and 1.30 e− per Å2 per frame. The second image dataset was acquired on a Titan Krios G4 microscope operating at 300 kV in nanoprobe mode using EPU for automated data collection. Movie frames were recorded on a Flacon 4i direct electron detector after a Selectris X energy filter using a 10 e-V slit width in zero-loss mode. Images were acquired at nominal magnification of 270,000, which yielded a pixel size of 0.458 Å.

Image analysis

For the first dataset movie frames were aligned, dose-weighted in Warp (Tegunov and Cramer, 2019) together with CTF estimation and particle picking. For the second dataset frame alignment, dose-weighting and CTF estimation was done in cryoSPARC (Punjani et al., 2017), picking was done with crYOLO (Wagner et al., 2019). These datasets were analyzed in cryoSPARC and RELION 3.1 (Zivanov et al., 2018) according to standard protocols. Briefly, images were subjected to reference-free 2D classification in cryoSPARC to remove images with contaminations. Ab-initio reconstruction was performed on the selected images followed by heterogenous and homogenous refinements in cryoSPARC. Structural variability and flexibility were explored with Relion 3.1 using 3D classification without alignment with different T values and cryoDRGN (Zhong et al., 2021) according to standard protocols. Sharpening of the maps was done with DeepEmhancer (Sanchez-Garcia et al., 2021).

Model building and refinement

A composite map was created from the part corresponding to the SIRT6 Rossmann-fold from the structure of the first dataset and the SIRT6 Zn-finger plus nucleosome from the structure of the second dataset. Crystal structure of Widom-601 sequence containing nucleosome (PDB: 3LZ0) and human SIRT6 (PDB: 5X16) were rigid-body fitted into this composite cryo-EM structure in Chimera (Pettersen et al., 2004). Flexible-fittings were performed with Isolde (Croll, 2018) in ChimeraX (Goddard et al., 2018) into the cryoDRGN structures. All models were refined in real-space with PHENIX (Liebschner et al., 2019) and Isolde.

Depiction of SIRT6 – nucleosome structures

The following maps are depicted in the figures:

Figure 1: EMD-16845; Figure 2: b1-cryoDRGN run1 PC1 volume 0, b2-volume 9; e1-cryoDRGN run2 PC1 volume 0, e2-volume 9; Figure 3: a,d-EMD-16845, e-EMD-18497; Figure 4: a-cryoDRGN run2 PC1 volume 0, b-volume 9; Supplemental Figure 4: a,b-EMD-16842, c-EMD-16843, d-EMD-18497; Supplemental Figure 5: a-EMD-16842, c-cryoDRGN run1 PC1 volume 0 and nucleosome void of SIRT6 (Suppl. Fig. 2), d-nucleosome void of SIRT6; Supplemental Figure 9: EMD-18497.

Molecular Dynamics simulations

All molecular dynamics (MD) simulations were performed with the AMBER20 suit of programs, using the ff14SB (Maier et al., 2015) force field with bsc1 DNA corrections (Ivani et al., 2016) and CUFIX ions corrections (Yoo and Aksimentiev, 2018) that improve the description of disordered region conformations. Parameters for the zinc finger were taken from Macchiagodena et al. (Macchiagodena et al., 2020).

The histone tails were added to the cryo-EM structure based on the work described by Armeev et al (Armeev et al., 2021), and protonation states of the protein residues were checked with the propka3 software (Olsson et al., 2011). The system was soaked in a truncated octrahedral TIP3P water box of 20 Å buffer and 0.15 mM of NaCl ions were added to simulate physiological conditions, resulting in a total of 330,000 atoms.

The minimization of the energy was carried out in 5 steps, each one alternated with a 20ps NVT equilibration step at 100K to ensure a very smooth optimization of the starting geometry. Each minimization run included 5,000 steps in steepest descent and 5,000 steps in conjugate gradient, with decreasing restraints on the protein and nucleic atoms from 20kcal/mol to 5kcal/mol that were kept in the associated equilibration step. The last minimization step was released of any constraints and followed by a final NVT equilibration run at 300K. The time step was then switched from 2 to 4fs as allowed by the use of the SHAKE bond length constraints (Kräutler et al., 2001) and the Hydrogen Mass Repartioning algorithm (Hopkins et al., 2015). The latter was applied on the topology using the parmed tool of Amber20. A 40ns equilibration run in NTP was then carried out, followed by a 5µs production. Coordinates were outputted every 1ns during the production run. The temperature was kept constant using the Langevin thermostat with a 2ps-1 collision frequency, and the Berendsen barostat was used to maintain the pressure at 1 atm in the NTP runs. Electrostatic interactions were treated using the Particle Mesh Ewald approach (Darden et al., 1993) with an 8 Å cut-off. Three replicates were performed starting with different velocities at the unconstrained 300K NVT equilibration step, resulting in a total of 15µs of sampling.

A cluster analysis was performed on the MD ensembles based on the DNA and protein RMSD. Conformations were extracted every 5 frames of the trajectories to avoid memory overload and were sorted using the hierarchical agglomerative and average linkage algorithms. 10 clusters were generated by replicate (Supplemental Table 2).

The DNA flexibility analysis was performed on each MD ensemble by using a Principal Component Analysis-based machine learning script quantifying the contribution of each residue to the overall dynamics of the system. Based on the Scikit-learn58 library (Fleetwood et al., 2020), it calculates a covariance matrix from the internal coordinates of the MD ensemble. In this matrix, the eigenmodes showing the highest amplitude are the most implicated in the main fluctuations of the system, and the contribution of each residue is normalized based on these values.

The control system was generated by removing SIRT6 from the initial structure and was submitted to the same protocol.

Data availability

The cryo-EM maps have been deposited in the Electron Microscopy Data Bank (EMDB) under accession codes EMD-16842 (High-resolution on the nucleosome), EMD-16843 (SIRT6-nucleosome refined for SIRT6), EMD-18497 (Tracing the H2A tail) and EMD-16845 (composite map). CryoDRGN maps are available upon request. The model coordinates for SIRT6-nucleosome were deposited in the Protein Data Bank (PDB) under the accession code 8OF4.

All MD input files are publicly available on Github: https://github.com/emmanuellebignon/NCP-SIRT6

Supporting information